---
title: "Objective3"
output: html_document
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This file explores Objective 3 of the project: To determine whether light levels and ships impact breeding birds and subsequently, chick provisioning behaviour. The file contains processing, analyses and plotting used for the final project write up.

## **Housekeeping**

```{r Install and Load packages}

#General
require(hms) #Time zone and handling times
require(dplyr) 
require(dbplyr)
require(lubridate)
require(tidyverse)
require(suncalc) #Calculates sun and moon elevation, rise and set
require(stringr)

#Plots
require(ggplot2) 
require(patchwork) #Compiling plots
require(ggeffects)
require(multcompView)

#Models
require(glmmTMB)
require(MASS) #Predictions of linear models (glm.nb)
require(lme4) #Fit linear and generalised linear mixed models (glmer)
require(ggfortify) #Plotting of statistics/inspecting residuals
require(car)
require(emmeans)
require(statmod)
require(betareg) #Beta regression

#Geolocator processing 
require(GeoLight)
require(TwGeos)
require(berryFunctions)
require(strucchange)
require(xfun)
require(data.table)
require(zoo)

#Mixture Model
require(plotmm)
require(mixtools)
```

## Hypothesis 6

Colony attendance of breeding shearwaters is significantly lower on nights with brighter skies.

#### Step 1: Read in Geolocator files in a usable format ('dark' 'light' 'wet' 'dry')

Due to the data used, I processed two separate read-ins: high resolution and low resolution. They require different code because I reduce the high resolution data down to fit the low resolution so that I can merge them.

Geolocator files are .deg (wet/dry data) and .lux (light/dark data). These both need to be read-in and merged with one another so that I can later find colony visits in the geolocator data.

```{r Low Resolution Read-In}
########## Low resolution data read-in 
########## (1) Reading in wet files first (.deg files)
#Set working directory where GLS files are found 
datadir <- setwd("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low")

#List all the files in the directory 
all_files <- list.files(path = datadir)

#Filter for only the .deg files in this directory (these are the appropriate immersion files for my loggers)
wet_files <- all_files[
  tolower(file_ext(all_files)) == "deg" &  #Must end with .deg
    !grepl("driftadj", all_files, ignore.case = TRUE)  #Exclude if "driftadj" appears anywhere
]

#Split the filename by the '_' character so we can add metadata to the GLS trace
#The way the lab group name files so that all metadata is included in it
splits <- strsplit(wet_files, "_")

#Create list of parameters for each file 
parameter <- list()
parameter$colony <- sapply(splits, "[[", 1)
parameter$ring <- sapply(splits, "[[", 2)
parameter$burrow <- sapply(splits, "[[", 3) %>% 
  str_remove_all("^b") %>% #Removing b before burrow name
  str_replace("^GB0+", "GB") #Removing 0 in GB01
parameter$gls <- sapply(splits, "[[", 4)
parameter$type <- sapply(splits, "[[", 5)

#Creating parameter for the dates is a bit more complicated
dates <- sapply(splits, "[", 6) %>%
  str_remove_all('.deg') #Removing file name extension .deg to leave just the date

#Creating characters for year, month and day
yrs <- NA
mns <- NA
dys <- NA

#Get the filename date into a usable format
for (i in 1:length(dates)) {
  if (nchar(dates[i]) == 8) {
    
    #Is year first?
    if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) != "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      #Is year second?
    } else if (substr(dates[i],1,2) != "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],5,8) 
      mn <- substr(dates[i],3,4)
      dy <- substr(dates[i],1,2) 
      
      #Is it hard to tell?
    } else if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- substr(dates[i],5,8) 
        mn <- substr(dates[i],3,4)
        dy <- substr(dates[i],1,2)  
      }
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- NA
        mn <- NA
        dy <- NA
      }
      
    }
    
  } else if (nchar(dates[i]) == 11) {
    yr <- NA
    mn <- NA
    dy <- NA
  }
  
  yrs[i] <- yr
  mns[i] <- mn
  dys[i] <- dy
  
}
parameter$year <- yrs
parameter$month <- mns
parameter$day <- dys

#Create a data frame for these parameters
parameter <- data.frame(parameter)

#Create a list to collect these parameter data frames for each file
coord_files <- list()

##Read-in loop
for (i in seq_along(wet_files)) {
  
  #Read in the files
  #z <- read.csv(paste(datadir, wet_files[i], sep = ''), skip = 19, sep = '')
  z <- tryCatch(
    read.csv(file.path(datadir, wet_files[i]), skip = 19, sep = '', stringsAsFactors = FALSE),
    error = function(e) {
      warning(sprintf("Failed to read %s: %s", wet_files[i], e$message))
      return(NULL)
    }
  )
  
  #Skip if read failed
  if (is.null(z)) next
  
  #Skip if zero rows
  if (nrow(z) == 0) {
    warning(sprintf("File %s read OK but has 0 rows — skipping.", wet_files[i]))
    next
  }
  
  #Combine date and time column
  z$DD.MM.YYYY <- paste(z$DD.MM.YYYY,z$HH.MM.SS)
  
  #Remove column 2 (the time column is no longer required)
  z[2] <- NULL
  
  #Rename columns to "POSIXct" (datetime) and (immersion) "score"
  names(z)[1] <- "POSIXct"
  names(z)[2] <- "score"
  
  #Convert date column to a time object (POSIXct format)
  z$POSIXct <- as.POSIXct(z$POSIXct, format = "%d/%m/%Y %H:%M:%OS", 'GMT')
  
  #Store this information as the first item in the list
  coord_files[[i]] <- z
  
  #Add columns for colony, ring, burrow, gls and file_year (all from the filename metadata), and also for season (taken as the year section of the POSIXct column)
  coord_files[[i]]$colony <- parameter$colony[i]
  coord_files[[i]]$ring <- parameter$ring[i]
  coord_files[[i]]$burrow <- parameter$burrow[i]
  coord_files[[i]]$gls <- str_remove_all(parameter$gls[i], "gls")
  coord_files[[i]]$file_year <- parameter$year[i]
  coord_files[[i]]$season <- format(coord_files[[i]]$POSIXct, "%Y")
  
  #Check that this has worked by printing each iteration number and the number of columns
  print(paste(i,ncol(coord_files[[i]])))
}

#Bind this list into a single data table
wetdatalowres <- rbindlist(coord_files, fill = TRUE)

#Create identifier title based on year and ring number?
wetdatalowres$title <- paste(wetdatalowres$ring, wetdatalowres$season, sep = ".")

#Save this data as an R object
write.csv(wetdatalowres, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/wetdatalow.csv")

########## (2) Read in light data files 
#Filter for only the .lux files in this directory
light_files <- all_files[
  tolower(file_ext(all_files)) == "lux" &  # must end with .lux
    !grepl("driftadj", all_files, ignore.case = TRUE)  # exclude if "driftadj" appears anywhere
]

##Split the filename by the '_' character so we can add metadata to the GLS trace
splits <- strsplit(light_files, 
                   "_")

#Create list of parameters for each file 
parameter <- list()
parameter$colony <- sapply(splits, "[[", 1)
parameter$ring <- sapply(splits, "[[", 2)
parameter$burrow <- sapply(splits, "[[", 3)
parameter$gls <- sapply(splits, "[[", 4)
parameter$type <- sapply(splits, "[[", 5)

#The "b" part of the burrow split may need to be removed - and also the 0 from GB01
parameter$burrow <- str_remove_all(parameter$burrow, '^b')
parameter$burrow <- stringr::str_replace(parameter$burrow, "^GB0+", "GB")

#Creating parameter for the dates is a bit more complicated
dates <- sapply(splits, "[", 6)

#Removing the file name extension from the date parameter
dates <- str_remove_all(dates, '.lux')
substr(dates,1,2)

yrs <- NA
mns <- NA
dys <- NA

#Get the filename date into a usable format
for (i in 1:length(dates)) {
  if (nchar(dates[i]) == 8) {
    
    # is the year first?
    if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) != "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      # is year second?
    } else if (substr(dates[i],1,2) != "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],5,8) 
      mn <- substr(dates[i],3,4)
      dy <- substr(dates[i],1,2) 
      
      # is it hard to tell?
    } else if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- substr(dates[i],5,8) 
        mn <- substr(dates[i],3,4)
        dy <- substr(dates[i],1,2)  
      }
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- NA
        mn <- NA
        dy <- NA
      }
      
    }
    
  } else if (nchar(dates[i]) == 11) {
    yr <- NA
    mn <- NA
    dy <- NA
  }
  
  yrs[i] <- yr
  mns[i] <- mn
  dys[i] <- dy
}

parameter$year <- yrs
parameter$month <- mns
parameter$day <- dys

#Create a data frame for these parameters
parameter <- data.frame(parameter)

#Create a list to collect these parameter data frames for each file
coord_files <- list()

#Read-in loop
for (i in seq_along(light_files)) {
  
  # read in the files (using the readMTlux function from the TwGeos package)  #install.packages("devtools") devtools::install_github("slisovski/TwGeos")
  z <- readMTlux(file.path(datadir, light_files[i]))

  # omit any NAs
  z <- na.omit(z)
  
  # rename the date/time column to "POSIXct"
  names(z)[1] <- "POSIXct"
  
  # store this information as the first item in the list
  coord_files[[i]] <- z
  
  # add metadata from the filename
  coord_files[[i]]$colony <- parameter$colony[i]
  coord_files[[i]]$ring <- parameter$ring[i]
  coord_files[[i]]$burrow <- parameter$burrow[i]
  coord_files[[i]]$gls <- str_remove_all(parameter$gls[i], "gls")
  coord_files[[i]]$file_year <- parameter$year[i]
  coord_files[[i]]$season <- format(coord_files[[i]]$POSIXct, "%Y")
  
  # check that this has worked by printing each iteration number and the number of columns
  print(paste(i, ncol(coord_files[[i]])))
}

#Bind this list into a single data table
lightdata <- rbindlist(coord_files, fill = TRUE)

#Save this as an R object
write.csv(lightdata, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/lightdatalow.csv")

########## (3) Merge wet and light data 
#Read in data files 
lightlow <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/lightdatalow.csv")
wetlow <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/wetdatalow.csv")

#Check colnames
colnames(lightlow)
colnames(wetlow)

#Make sure colnames are the same and datetime in POSIXct 
wetlow$POSIXct <- as.POSIXct(wetlow$POSIXct, format = "%Y-%m-%d %H:%M:%S")
lightlow$POSIXct <- as.POSIXct(lightlow$POSIXct, format = "%Y-%m-%d %H:%M:%S")

#Merge based on title, POSIXct, colony, file_year, 
lightwet.datalow <- merge(lightlow, wetlow, by = c("POSIXct", "colony", "ring", "burrow", "gls", "file_year"), all = T)

#Can check that this does indeed contain Light and immersion (score) data
hist(lightwet.datalow$Light)
hist(lightwet.datalow$score)

#Check that all the rings are present (there should be 15)
unique(lightwet.datalow$ring)

#Save this 
write.csv(lightwet.datalow, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/lightwet_datalow.csv")
```

```{r High Resolution Read-In}
########## High Resolution 
########## (1) Wet data 
datadirhigh <- setwd("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high")

wet_files_high <- list.files(path = datadirhigh) %>%
  tolower() %>% 
  .[file_ext(.) == "deg" & !grepl("driftadj", ., ignore.case = TRUE)] #Must end with .deg and exclude if "driftadj" appears anywhere

splitshigh <- strsplit(wet_files_high, "_")

parameterhigh<-list()
parameterhigh$colony <- sapply(splitshigh, "[[", 1)
parameterhigh$ring <- sapply(splitshigh, "[[", 2)
parameterhigh$burrow <- sapply(splitshigh, "[[", 3) %>% 
  str_remove_all("^b") %>% #Removing b before burrow name
  str_replace("^GB0+", "GB") #Removing 0 in GB01
parameterhigh$gls <- sapply(splitshigh, "[[", 4)
parameterhigh$type <- sapply(splitshigh, "[[", 5)

#Creating parameter for the dates is a bit more complicated
dateshigh <- sapply(splitshigh, "[", 6) %>%
  str_remove_all('.deg') #Removing file name extension .deg to leave just the date

#Creating characters for year, month and day
yrs <- NA
mns <- NA
dys <- NA

#Get the filename date into a usable format
for (i in 1:length(dateshigh)) {
  if (nchar(dateshigh[i]) == 8) {
    
    #Is year first?
    if (substr(dateshigh[i],1,2) == "20" & substr(dateshigh[i],5,6) != "20") {
      yr <- substr(dateshigh[i],1,4) 
      mn <- substr(dateshigh[i],5,6)
      dy <- substr(dateshigh[i],7,8) 
      
      #Is year second?
    } else if (substr(dateshigh[i],1,2) != "20" & substr(dateshigh[i],5,6) == "20") {
      yr <- substr(dateshigh[i],5,8) 
      mn <- substr(dateshigh[i],3,4)
      dy <- substr(dateshigh[i],1,2) 
      
      #Is it hard to tell?
    } else if (substr(dateshigh[i],1,2) == "20" & substr(dateshigh[i],5,6) == "20") {
      yr <- substr(dateshigh[i],1,4) 
      mn <- substr(dateshigh[i],5,6)
      dy <- substr(dateshigh[i],7,8) 
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- substr(dateshigh[i],5,8) 
        mn <- substr(dateshigh[i],3,4)
        dy <- substr(dateshigh[i],1,2)  
      }
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- NA
        mn <- NA
        dy <- NA
      }
      
    }
    
  } else if (nchar(dateshigh[i]) == 11) {
    yr <- NA
    mn <- NA
    dy <- NA
  }
  
  yrs[i] <- yr
  mns[i] <- mn
  dys[i] <- dy
  
}

parameterhigh$year <- yrs
parameterhigh$month <- mns
parameterhigh$day <- dys

#Create a data frame for these parameters
parameterhigh <- data.frame(parameterhigh)

#Create a list to collect these parameter data frames for each file
coord_files_high <- list()

#VIN2RIN function needed for high res read in 
vin2rin <- function(input){
temp_df <-zhigh[which(is.na(zhigh$big.number) == F),]
zhigh$big.number <- as.character(zhigh$big.number)
zhigh$big.number <- as.numeric(zhigh$big.number)
reversal <- data.frame(zhigh$big.number, temp_df$state)
expansion <- reversal[rep(row.names(reversal),reversal$zhigh.big.number), 2]

chunks <- split(expansion, ceiling(seq_along(expansion)/300))

wet_freq <- plyr::ldply(chunks, function(c) sum(c=="wet"))
dry_freq <- plyr::ldply(chunks, function(c) sum(c=="dry"))

#Check to determine start time direction
forwardStt <- temp_df$POSIXct[1]
forwardCheck <- temp_df$POSIXct[1] + zhigh$big.number[1]
backwardStt <- temp_df$POSIXct[1] - zhigh$big.number[1]
forwardDate <- strsplit(as.character(forwardCheck), split = " ")
forwardDate <- forwardDate[[1]][1]

sttime <- ifelse(forwardDate == as.Date(temp_df$POSIXct[[2]]), as.character(forwardStt), as.character(backwardStt))
sttime <- as.POSIXct(sttime, format = "%Y-%m-%d %H:%M:%S", 'GMT')

forwardendtime <- tail(temp_df$POSIXct, n =1)+tail(zhigh$big.number, n = 1)
backendtime <- tail(temp_df$POSIXct, n =1)

endtime <- ifelse(forwardDate == as.Date(temp_df$POSIXct[[2]]), as.character(forwardendtime), as.character(backendtime))
endtime <- as.POSIXct(endtime, format = "%Y-%m-%d %H:%M:%S", 'GMT')

outlength <- ceiling(sum(zhigh$big.number)/300)
POSIXct <- seq.POSIXt(sttime, by = "300 secs", length.out = outlength)
POSIXct <- as.data.frame(POSIXct)

dry <- dry_freq$V1
wet <- wet_freq$V1

minbins <- as.data.frame(cbind(POSIXct, wet, dry))

minbins <- minbins[-nrow(minbins),]  

fullset <- minbins
fullset$length <- fullset$wet/3
fullset$length <- round(fullset$length)
fullset$dry <- NULL
fullset$wet <- NULL
fullset$status <- 'HIGHRES'
fullset$exceltime <- NA

colnames(fullset)[1] <- 'datetime'
return(fullset)
}

#Read-in loop
for (i in seq_along(wet_files_high)) {
  
  # read in the files
  zhigh <- tryCatch(
    read.csv(file.path(datadirhigh, wet_files_high[i]), skip = 19, sep = '', stringsAsFactors = FALSE),
    error = function(e) {
      warning(sprintf("Failed to read %s: %s", wet_files_high[i], e$message))
      return(NULL)
    }
  )
    
    names(zhigh)[1] <- "POSIXct"
    names(zhigh)[2] <- "time"
    names(zhigh)[3] <- "big.number"
    names(zhigh)[4] <- "state"
    #rename columns
    #if already summed into ten min bins
   
     #Combine date and time column
    zhigh$POSIXct <- paste(zhigh$POSIXct,zhigh$time)
    
    # remove column 2 (the time column is no longer required)
    zhigh[2] <- NULL
    
    # convert date column to a time object (POSIXct format)
    zhigh$POSIXct <- as.POSIXct(zhigh$POSIXct, format = "%d/%m/%Y %H:%M:%OS", 'GMT')
  
    zhigh<-vin2rin(zhigh)
    
    #convert date column to a time object specifying the input format or it gets confused
    coord_files_high[[i]] <- zhigh
  
  # add metadata from the filename
  coord_files_high[[i]]$colony <- parameterhigh$colony[i]
  coord_files_high[[i]]$ring <- parameterhigh$ring[i]
  coord_files_high[[i]]$burrow <- parameterhigh$burrow[i]
  coord_files_high[[i]]$file_year <- parameterhigh$year[i]
  coord_files_high[[i]]$gls <- str_remove_all(parameterhigh$gls[i], "gls")

  # add an identifier
  coord_files_high[[i]]$trackID <- i
  
  print(paste(i,ncol(coord_files_high[[i]])))
}

#Bind this list into a single data table
wetdatahighres <- rbindlist(coord_files_high, fill = TRUE)

#Create identifier title based on year and ring number?
wetdatahighres$title <- paste(wetdatahighres$ring, wetdatahighres$file_year, sep = ".")

#Save this data as an R object
write.csv(wetdatahighres, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/wetdatahigh.csv")

########## (2) Light data 
all_files <- list.files(path = datadirhigh)

#Filter for only the .lux files
light_files_high <- all_files[
  tolower(file_ext(all_files)) == "lux" &  # must end with .lux
    !grepl("driftadj", all_files, ignore.case = TRUE)  # exclude if "driftadj" appears anywhere
]

##Split the filename by the '_' character so we can add metadata to the GLS trace
splits <- strsplit(light_files, 
                   "_")

#Create list of parameters for each file 
parameter <- list()
parameter$colony <- sapply(splits, "[[", 1)
parameter$ring <- sapply(splits, "[[", 2)
parameter$burrow <- sapply(splits, "[[", 3)
parameter$gls <- sapply(splits, "[[", 4)
parameter$type <- sapply(splits, "[[", 5)

#The "b" part of the burrow split may need to be removed - and also the 0 from GB01
parameter$burrow <- str_remove_all(parameter$burrow, '^b')
parameter$burrow <- stringr::str_replace(parameter$burrow, "^GB0+", "GB")

#Creating parameter for the dates is a bit more complicated
dates <- sapply(splits, "[", 6)

#Removing the file name extension from the date parameter
dates <- str_remove_all(dates, '.lux')
substr(dates,1,2)

yrs <- NA
mns <- NA
dys <- NA

#Get the filename date into a usable format
for (i in 1:length(dates)) {
  if (nchar(dates[i]) == 8) {
    
    # is the year first?
    if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) != "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      # is year second?
    } else if (substr(dates[i],1,2) != "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],5,8) 
      mn <- substr(dates[i],3,4)
      dy <- substr(dates[i],1,2) 
      
      # is it hard to tell?
    } else if (substr(dates[i],1,2) == "20" & substr(dates[i],5,6) == "20") {
      yr <- substr(dates[i],1,4) 
      mn <- substr(dates[i],5,6)
      dy <- substr(dates[i],7,8) 
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- substr(dates[i],5,8) 
        mn <- substr(dates[i],3,4)
        dy <- substr(dates[i],1,2)  
      }
      
      if (as.numeric(mn) < 4 | as.numeric(mn) > 10) {
        yr <- NA
        mn <- NA
        dy <- NA
      }
      
    }
    
  } else if (nchar(dates[i]) == 11) {
    yr <- NA
    mn <- NA
    dy <- NA
  }
  
  yrs[i] <- yr
  mns[i] <- mn
  dys[i] <- dy
}

parameter$year <- yrs
parameter$month <- mns
parameter$day <- dys

#Create a data frame for these parameters
parameter <- data.frame(parameter)

#Create a list to collect these parameter data frames for each file
coord_files <- list()

#Read-in loop!
for (i in seq_along(light_files)) {
  
  # read in the files (using the readMTlux function from the TwGeos package)  #install.packages("devtools") devtools::install_github("slisovski/TwGeos")
  z <- readMTlux(file.path(datadir, light_files[i]))
  
  # omit any NAs
  z <- na.omit(z)
  
  # rename the date/time column to "POSIXct"
  names(z)[1] <- "POSIXct"
  
  # store this information as the first item in the list
  coord_files[[i]] <- z
  
  # add metadata from the filename
  coord_files[[i]]$colony <- parameter$colony[i]
  coord_files[[i]]$ring <- parameter$ring[i]
  coord_files[[i]]$burrow <- parameter$burrow[i]
  coord_files[[i]]$gls <- str_remove_all(parameter$gls[i], "gls")
  coord_files[[i]]$file_year <- parameter$year[i]
  coord_files[[i]]$season <- format(coord_files[[i]]$POSIXct, "%Y")
  
  # check that this has worked by printing each iteration number and the number of columns
  print(paste(i, ncol(coord_files[[i]])))
}

#Bind this list into a single data table
lightdata <- rbindlist(coord_files, fill = TRUE)

#Create identifier title based on year and ring number
lightdata$title <- paste(lightdata$ring, lightdata$season, sep = ".")

#Save this as an csv
write.csv(lightdata, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/lightdatahigh.csv")

########## (3) Merge wet and light data
#Read in files created above
wet <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/wetdatahigh.csv")
light <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/lightdatahigh.csv")

#Make sure they have the same col names
colnames(wet) 
colnames(light)

#Ensure POSIXct in same format 
#Burrow names and ring nummbers are lower and upper case - need to make consistent 
wet <- wet %>%
  mutate(
    POSIXct = as.POSIXct(datetime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"),
    ring    = recode(ring,
                     em35000 = "EM35000",
                     em55662 = "EM55662",
                     ea05370 = "EA05370"),
    burrow  = recode(burrow,
                     `16l` = "16L",
                     hg16  = "HG16"),
    gls     = recode(gls,
                     cl389 = "CL389",
                     cl407 = "CL407",
                     ce989 = "CE989"))

light <- light %>%
  mutate(
    POSIXct = as.POSIXct(POSIXct, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))

lightwet.datahigh <- merge(light, wet, by = c("POSIXct", "ring", "burrow", "file_year", "gls", "colony"), all = T)

#Can check that this does indeed contain Light and immersion (score) data
hist(lightwet.datahigh$Light)
hist(lightwet.datahigh$length)

#Check that all the rings are present that you intend 
unique(lightwet.datahigh$ring)

#Save this so light and wet data are all together 
write.csv(lightwet.datahigh, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/lightwet_datahigh.csv")
```

```{r Merge High Resolution and Low Resolution Data}
#Get both files saved above: low res wet and light, high res wet and light
lowres <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/low/lightwet_datalow.csv")
highres <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/high/lightwet_datahigh.csv")

#Check ring numbers in each
unique(lowres$ring)
unique(highres$ring)

#Check colnames - are they the EXACT same 
colnames(lowres)
colnames(highres)

#Remove columns I dont need and adding ones so they match
lowres <- lowres[ , !(names(lowres) %in% c("X", "X.x", "season.y", "X.y")) ]
highres <- highres[ , !(names(highres) %in% c("X.1", "X", "exceltime", "datetime", "trackID", "n_obs", "lux_median", "lux_sum", "X.y", "X.x")) ]
lowres$status <- "LOWRES"
highres$season <- highres$file_year

#Renaming columns so they are the same (need matching for row binding)
names(highres)[names(highres) == "Light"] <- "lux"
names(highres)[names(highres) == "length"]   <- "wet_score"
names(lowres)[names(lowres) == "Light"]   <- "lux"
names(lowres)[names(lowres) == "score"]   <- "wet_score"
names(lowres)[names(lowres) == "season.x"]   <- "season"

#Combine
all <- rbind(lowres, highres)

#Save to directory 
write.csv(all, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/all.csv")
```

#### Step 2: Classify Colony Visitation ('dark' and 'dry' periods)

```{r Housekeeping for Geolocator Data}
#Load GLS data
data <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/Processed Files/all.csv")

#Inspect
hist(data$lux)
hist(data$wet_score)
plot(data$lux, data$wet_score)

#Remove geolocator data when the loggers were not on the birds
#Period between when data was taken from from bird and when it was downloaded is removed

#Get dates when Geolocators were on (fieldwork book)
GLSdiary <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/GLSdiary.csv")

#Use POSIXct for dates and times
#Deploy = put on bird, Retrieval = taken off bird
GLSdiary$Deploy.POSIXct <- paste(GLSdiary$Attached.Date, GLSdiary$Attached.time..approx.)
GLSdiary$Deploy.POSIXct <- as.POSIXct(GLSdiary$Deploy.POSIXct, format = "%d/%m/%Y %H:%M:%S", tz = "UTC")

GLSdiary$Retrieval.POSIXct <- paste(GLSdiary$Retrieval.Date.from.bird, GLSdiary$Retrieval.time.from.bird)
GLSdiary$Retrieval.POSIXct <- as.POSIXct(GLSdiary$Retrieval.POSIXct, format = "%d/%m/%Y %H:%M:%S", tz = "UTC")

#Are Deploy/Retrieval times missing?
table(is.na(GLSdiary$Deploy.POSIXct))
table(is.na(GLSdiary$Retrieval.POSIXct))

#Join to the GLS wet and light data based on ring number (excat deployment and retrieval for each bird)
#First need to change name of ring to same to match them
names(GLSdiary)[names(GLSdiary) == "Ring"] <- "ring"

data<- data %>%
  left_join(
    GLSdiary %>%
      select(Deploy.POSIXct, Retrieval.POSIXct, ring),
    by = "ring")

#Check after your left_join, how many rows lost per ring?
joined_summary <- data %>% group_by(ring) %>%
  summarize(n_rows = n(),
            n_na_deploy = sum(is.na(Deploy.POSIXct)),
            n_na_retrieval = sum(is.na(Retrieval.POSIXct)))
print(joined_summary)

#Rings in your joined_summary with all NA deploy/retrieval
bad_rings <- joined_summary %>% filter(n_na_deploy == n_rows | n_na_retrieval == n_rows) %>% pull(ring)
bad_rings

#Show a quick count
length(bad_rings)

#Remove rows that are before the last deployment and after the first retrieval
#To ensure the same sample size for whole duration
#Last deployment was 2025-07-15 at 1:15am (AF47, EA05386)
#First retrieval was 2025-08-08 01:43am (426, EM91278)
data <- data %>%
  filter(
    POSIXct >= "2025-07-15 01:15:00" &
      POSIXct <= "2025-08-08 01:43:00")

#Only keep dates within deployment and retrieval 
data <- data %>%
  filter(POSIXct >= Deploy.POSIXct & POSIXct <= Retrieval.POSIXct)

#Check how many unique rings before and after filtering
#See how many are lost and why 
length(unique(read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/Processed Files/all.csv")$ring))
length(unique(data$ring))

#6 rings lost only - This is fine because they are outside when the Sky Quality Meter was deployed so not used in this study
#"EA59292" "EJ99646" "EM55664" "EM55768" "EM55771" "EY07308"
length(unique(GLSdiary$ring))
```

```{r Extract Dry and Wet, Light and Dark}
#Extract dates we have geolocators on for
skomerdates<- as.Date(format(as.POSIXct(data$POSIXct, "%Y-%m-%d", tz = "UTC") ) )

#Get sunset and sunrise for these dates
sun_skomer<-getSunlightTimes(date=skomerdates, lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset"))
data <- cbind(data, sun_skomer)

#Make sure all columns are POSIXct format
data$sunrise  <- as.POSIXct(data$sunrise, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
data$sunset   <- as.POSIXct(data$sunset, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
data$POSIXct <- as.POSIXct(data$POSIXct, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

#Check any NA sunrise/sunset now:
cat("NA sunrise rows:", sum(is.na(data$sunrise)), "NA sunset rows:", sum(is.na(data$sunset)), "\n")

#Get rid of columns don't need 
data <- subset(data, select = -c(date, X, lat, lon))

#Define day and night
data <- data %>%
  mutate(daynight = case_when(
    is.na(sunrise) | is.na(sunset) ~ NA_character_,
    POSIXct >= sunrise & POSIXct <= sunset ~ "Day",
    TRUE ~ "Night"))

#Check counts per ring and per day/night
summary_table <- data %>%
  group_by(ring, daynight) %>%
  summarise(n_rows = n()) %>% ungroup()
print(head(summary_table, 20))

#How many rings have ANY Night rows? ANY Day rows? Should have 47 rings still!
rings_with_night <- summary_table %>% filter(daynight=="Night") %>% pull(ring) %>% unique()
rings_with_day   <- summary_table %>% filter(daynight=="Day")   %>% pull(ring) %>% unique()
cat("Rings with Night rows:", length(rings_with_night), "  Rings with Day rows:", length(rings_with_day), "\n")

#Create column of numbers for changes in state from day/night to create an ID number
data<-  data%>%
  mutate(
    change = case_when(
      daynight != lag(daynight) ~ TRUE,
      TRUE ~ FALSE),
    n_change = cumsum(change))

#Create ID for night and day number 
data$daynight.id<- paste(data$daynight, data$n_change, sep=".")

#Calculate the time difference between rows so we can sum times for mixture model (for when we have data of lots of different time periods)
data<- split(data, data$ring) #Split data so that the loop runs on every ring number
for (i in 1:length(data)){
  infile = data[[i]] 
  
  infile<-infile %>%  
    mutate(timediff= c(0, difftime(infile$POSIXct[-length(infile$POSIXct)], infile$POSIXct[-1], unit = "mins"))) #Gives the time gap between successive rows in minutes
  infile$timediff[1]<- infile$timediff[2] #First row doesnt have a time diff so instead of being 0, it is the same as the second row 
  
  data[[i]]<- infile
  print(i) #Should have number same as number of rings you have
}

data<- bind_rows(data)
data$timediff<- abs(data$timediff)

#If any weird negatives, inspect:
if(any(data$timediff < 0, na.rm=TRUE)) cat("Warning: negative timediff present\n")

#Seperate nights from days (what we are interested in)
nights<- filter(data, data$daynight=="Night")

#Count number of bins with wet count - if score is more than 1 then wet, if not then dry
nights$wet<- with(nights, ifelse(wet_score >= 1 , "Wet", "Dry"))

#Sum the length of each night by adding the time different between rows 
countspernight<-aggregate(x = nights$timediff,                # Specify data column
                          by = list(nights$daynight.id),              # Specify group indicator
                          FUN = sum)

#Count number of wet or dry events per night
scorepernight<-nights%>%
  group_by(daynight.id, wet) %>% #Groups data by the night and also by wet or dry within that night
  select(wet)  %>%
  dplyr::count() #Counts how many dry or how many wet in that night 

#Join length of night and count data per night
#Creates a dry and wet column for each night ID so they are not duplicates
scorepernight <-scorepernight%>% 
  pivot_wider(names_from=wet, values_from=n)
#Where there is NA, put 0 (because NA is where there is no wet or dry count)
scorepernight[is.na(scorepernight)] <- 0

#Remove other columns to get wetcount only 
wetcountonly <- scorepernight[-c(1:2)]
names(wetcountonly)[1]<- "wetscore"
wetcountonly<- cbind(countspernight, wetcountonly)

#Plot a histogram 
par(mfrow=c(1,1))
hist(wetcountonly$wetscore, xlab=
       "Number of Wet Bins per Night", breaks=30, main="", col="Dark grey")
```

```{r Mixture Model}
#Run mixture model with EM algorithm to see which nights belong to whcih distribution 
#Two separate distributions - one for colony visits and one for short boughts of flight 

#Run mix model based on k=2 (Number of components = k, Default = 2)
m1<-normalmixEM(wetcountonly$wetscore, k=2)
#Plot density output
par(mfrow=c(1,1))
plot(m1, which=2, breaks=30)

if(require(mixtools)){
  mixmdl <- mixtools::normalmixEM(wetcountonly$wetscore, k = 2)
}
x <- mixmdl$x
x <- data.frame(x)
ggplot2::ggplot(data.frame(x)) +
  ggplot2::geom_density(ggplot2::aes(x), color="dark grey", fill="dark grey") +
  ggplot2::stat_function(geom = "line", fun = plot_mix_comps_normal,
                         args = list(mixmdl$mu[1], mixmdl$sigma[1], lam = mixmdl$lambda[1]),
                         colour = "red", linewidth=2) +
  ggplot2::stat_function(geom = "line", fun = plot_mix_comps_normal,
                         args = list(mixmdl$mu[2], mixmdl$sigma[2], lam = mixmdl$lambda[2]),
                         colour = "blue", linewidth=2) +theme_classic()+ xlab("Number of Wet Bins per Night") +ylab("Density")

#mu = vector of component means - mean wet bins for comp 1 and 2
#sigma - vector component standard deviance 
#lambda - mixing proportions that sum to 1 (if 0.7, 0.3 then it means 70% of nights in component 1)

#Look at probability that a night belongs to each distribution
post.df <- as.data.frame(cbind(x = m1$x, m1$posterior))

names(post.df)[2]<- "Col"
names(post.df)[3]<- "No_Col"
post.df<- cbind(post.df, wetcountonly)
names(post.df)[4]<- "daynight.id"

nights<- split(nights, nights$daynight.id)

title<-post.df$daynight.id

#Make sure nights are in right order
post.df<- split(post.df, post.df$daynight.id)
title<-names(post.df)
nights<-nights[title]

for (i in 1 : length (nights)){
  infile<- nights[[i]]
  infile2<- post.df[[i]]
  infile$colprob<- infile2$Col
  nights[[i]]<- infile
  name = names(nights[i]) 
  names(nights)[i]<-name
  print(i)
}

#Add in col prob info
nights<- bind_rows(nights)

#Define day probability of having a colony visit
days<- filter(data, data$daynight=="Day")
lux_threshold <- 1.136
data <- data %>%
  mutate(wet_flag = ifelse(wet_score >= 1, TRUE, FALSE),
         state = case_when(
           daynight == "Day" & (wet_score == 0 & (lux <= lux_threshold)) ~ "Dark_dry",
           daynight == "Day" ~ "notofcurrentinterest",
           TRUE ~ NA_character_))

days<- split(days, days$daynight.id)
for(i in 1:length(days)){
  infile<- days[[i]]
  
  infile<- infile %>%
    mutate(
      change = case_when(
        state != lag(state) ~ TRUE,
        TRUE ~ FALSE
      ),
      n_change = cumsum(change)
    )                         
  
  check<-aggregate(x = infile$timediff,                # Specify data column
                   by = list(infile$state, infile$n_change),              # Specify group indicator
                   FUN = sum)
  
  if(max(check$x)>=360) {
    if(check$Group.1[which.max(check$x)]=="notofcurrentinterest"){
      infile$colprob=0
    }
    
    if(check$Group.1[which.max(check$x)]=="Dark_dry"){
      infile$colprob=1
    }
  }
  
  if(max(check$x)<=360){
    infile$colprob=0
  }
  
  
  days[[i]]<-infile
  
}

##look for long dark periods in the day
days<- bind_rows(days)
#days$colprob<- with(days, ifelse(state=="Dark_dry", 1, 0))
days$state<- NULL
days$nstate<- NULL
days$check<- NULL
nights$wet<-NULL
#put probability of colony visit at 1 if there is dark_dry during day light

#Combine days and nights
combined<- rbind(days, nights)

#Create dataset of colony visits
combined<- combined[order(combined$ring, combined$POSIXct),] 
#Reorder it so its grouped by ring and time

#Use a threshold to determine if bird is on the colony 
hist(combined$colprob) 
#This histogram shows that there are 2 main distributions - clear cut can be made between the two

#Recommend a threshold of 0.8 because my distributions don't overlap too much
#Increased certainty 
combined$colony.visit <- with(combined, ifelse(colprob >= 0.8, "Yes", "No"))

#Save this as a csv file 
write.csv(combined, file = "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/Processed files/colonyvisits.csv")
```

#### Step 3: Calculate the number of birds arriving each night

```{r Extracting number of birds for each night}
#Load gls data from above
glscount <- read.csv( "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/GLS/CountGLS.csv")

#Best nightly metric is how many arrived each night - sum up within a night 
glsnightly <- glscount %>%
  group_by(night.of) %>%
  summarise(
    n_arrived   = sum(n_individuals_arrived, na.rm = TRUE),
    mean_natural = mean(illumination, na.rm = TRUE),
    max_natural = max(illumination, na.rm = TRUE),
    min_natural = min(illumination, na.rm = TRUE),
    med_natural = median(illumination, na.rm = TRUE),
    med_logcdm2 = median(logcdm2, na.rm = TRUE), 
    mean_logcdm2 = mean(logcdm2, na.rm = TRUE),
    med_cdm2 = median(cdm2, na.rm = TRUE),
    mean_cdm2 = mean(cdm2, na.rm = TRUE),
    mean_moonPhase   = mean(moonPhase, na.rm = TRUE),
    .groups = "drop")

#Remove date 07/31 - not light data due to being off fieldwork that day 
glsnightly <- glsnightly[glsnightly$night.of != "2025-07-31", ]

#Combine with ship data too
ship <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Analysis/analysis/data/shipwide.csv")

#Adding back midnight time stamps and POSIXct
ship$POSIXct <- as.POSIXct(
  ifelse(nchar(as.character(ship$POSIXct)) == 10,
         paste(ship$POSIXct, "00:00:00"),
         as.character(ship$POSIXct)),
  format="%Y-%m-%d %H:%M:%S")

#Night.of id
sun <-getSunlightTimes(date = as.Date(ship$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset"))
ship <- cbind(ship, sun)
ship <- subset(ship, select = -c(date, lat, lon))
#Give night.id again (same as before)
ship$night.of <- with(ship, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

#Average the heading, distance, number of ships across the night
shipnightly <- ship %>%
  group_by(night.of) %>%
  summarise(
    avg_heading   = mean(avg_heading, na.rm = TRUE),
    med_heading = median(avg_heading, na.rm = TRUE),
    avg_distance = mean(avg_distance_m, na.rm = TRUE),
    med_distance = median(avg_distance_m, na.rm = TRUE),
    avg_ship_no = ceiling(mean(no.ships,na.rm = TRUE)),
    .groups = "drop")

#Bind to the rest of the data 
allnightly <- glsnightly %>%
  left_join(shipnightly, by = "night.of")

allnightly <- allnightly[!is.na(allnightly$avg_ship_no), ]
```

#### Step 4: Model Fitting

```{r Fixed and Random Effects Creation}
#Days since Start
allnightly$days_from_start <-as.numeric(as.Date(allnightly$night.of) - min(as.Date(allnightly$night.of), na.rm = TRUE))
```

```{r Model}
#Model with natural illumination
model_full <- glm.nb(n_arrived ~ avg_ship_no + med_logcdm2 + med_natural + days_from_start, data = allnightly)
summary(model_full)

#Model with only ships
model_reduced <- glm.nb(n_arrived ~ avg_ship_no + days_from_start, data = allnightly)
summary(model_reduced)
```

```{r Plot Full Model}
#Prediction for logcdm on full model
newdat_full <- expand.grid(
  med_logcdm2 = seq(min(allnightly$med_logcdm2, na.rm = TRUE),
                    max(allnightly$med_logcdm2, na.rm = TRUE),
                    length.out = 100),
  avg_ship_no = mean(allnightly$avg_ship_no), na.rm = TRUE,
  med_natural = mean(allnightly$med_natural), na.rm = TRUE,
  days_from_start = mean(allnightly$days_from_start), na.rm = TRUE)
pred_out2 <- predict(model_full, newdat_full, type = "link", se.fit = TRUE, re.form = NA)
newdat_full$fit_link <- pred_out2$fit
newdat_full$se_link  <- pred_out2$se.fit
#95% CI on link scale, then transform back
newdat_full$pred <- exp(newdat_full$fit_link)
newdat_full$lwr  <- exp(newdat_full$fit_link - 1.96 * newdat_full$se_link)
newdat_full$upr  <- exp(newdat_full$fit_link + 1.96 * newdat_full$se_link)
newdat_full <- newdat_full[, !duplicated(names(newdat_full))]

#Plot cdm2 on full model 
cdm <- ggplot(allnightly, aes(x = med_logcdm2, y = n_arrived)) +
  geom_point() +
  geom_line(data = newdat_full, aes(x = med_logcdm2, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = newdat_full,
              aes(x = med_logcdm2, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters Arrived",
    breaks = seq(0, max(allnightly$n_arrived, na.rm = TRUE), by = 5))+
  coord_cartesian(ylim = c(0, 18)) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
```

```{r Plot Reduced Model}
#Prediction for ships (reduced model)
newdat_reduced <- expand.grid(
  avg_ship_no = seq(min(allnightly$avg_ship_no, na.rm = TRUE),
                       max(allnightly$avg_ship_no, na.rm = TRUE),
                       length.out = 100),
  days_from_start = mean(allnightly$days_from_start), na.rm = TRUE)
pred_out <- predict(model_reduced, newdat_reduced, type = "link", se.fit = TRUE, re.form = NA)
newdat_reduced$fit_link <- pred_out$fit
newdat_reduced$se_link  <- pred_out$se.fit
#95% CI on link scale, then transform back
newdat_reduced$pred <- exp(newdat_reduced$fit_link)
newdat_reduced$lwr  <- exp(newdat_reduced$fit_link - 1.96 * newdat_reduced$se_link)
newdat_reduced$upr  <- exp(newdat_reduced$fit_link + 1.96 * newdat_reduced$se_link)

#Plot ships on reduced model
ship <- ggplot(allnightly, aes(x = avg_ship_no, y = n_arrived)) +
  geom_point() +
  geom_line(data = newdat_reduced, aes(x = avg_ship_no, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = newdat_reduced,
              aes(x = avg_ship_no, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters Arrived",
    breaks = seq(0, max(allnightly$n_arrived, na.rm = TRUE), by = 5))+
  coord_cartesian(ylim = c(0, 18)) +
  scale_x_continuous(
    name = expression("Number of Ships")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))

#Patching figures together 
(cdm|ship) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```

## Hypothesis 7

Chicks are less likely to be fed by adults on nights with brighter skies.

#### Step 1: Extract meal size from chick weights

Chicks were weighed once a day. We factor in metabolic loss into an estimation of nightly meal size, using the equation outlined in the project write-up.

```{r Extract meal size}
#Load Data
weights <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Chick Weights/Chickweights.csv")

#POSIXct the check date and time 
weights$POSIXct <- paste(weights$Check.date, weights$Check.time)
weights$POSIXct <- as.POSIXct(weights$POSIXct, tz = "UTC", "%d/%m/%Y %H:%M:%OS")

#Weight should be numeric 
weights$weight.g <- as.numeric(weights$Weight..g.)

#Remove columns not needed 
weights <- weights[, !(names(weights) %in% c("Weight..g.", "X", "X.1", "X.2", "X.3", "X.4", "X.5", "X.6", "X.7", "X.8", "X.9", "X.10", "X.11", "X.12", "X.13", "X.14", "X.15", "X.16", "X.17", "X.18", "X.19", "X.20"))]

#Remove NAs in POSIXct or Weight
weights <- weights[!is.na(weights$POSIXct) & !is.na(weights$weight.g), ]

#Apply equation of metabolic loss (r1 x time(hours)) (grouping by burrow and time)
#Coefficients and SEs
a_coef <- 2.21e-3
a_se   <- 0.61e-3
b_coef <- 1.50
b_se   <- 0.22

#Compute overnight metabolic loss to the next day's observation for each burrow
meal_size <- weights %>%
  arrange(Burrow, POSIXct) %>%
  group_by(Burrow) %>%
  mutate(
    next_POSIXct = lead(POSIXct),
    next_mass = lead(weight.g),
    hours   = as.numeric(difftime(next_POSIXct, POSIXct, units = "hours")),
    initial_mass = weight.g,
    r1 = a_coef * (initial_mass) + b_coef,
    metabolic_loss = r1 * hours, 
    var_a = a_se^2,
    var_b = b_se^2,
    se_loss = abs(hours) * sqrt( (initial_mass^2) * var_a + var_b ),
    ci_lower = metabolic_loss - 1.96 * se_loss,
    ci_upper = metabolic_loss + 1.96 * se_loss) %>%
  ungroup() %>%
  filter(!is.na(hours)) %>%
  select(Burrow, POSIXct, next_POSIXct, hours, initial_mass, next_mass,
         metabolic_loss, se_loss, ci_lower, ci_upper)

#Create a column called change in mass 
meal_size$change_mass <- meal_size$next_mass - meal_size$initial_mass

#Calculate meal size = change_mass + metabolic loss
meal_size$meal_size <- meal_size$change_mass + meal_size$metabolic_loss
meal_size$meal_lower <- meal_size$change_mass + meal_size$ci_lower
meal_size$meal_upper <- meal_size$change_mass + meal_size$ci_upper

```

#### Step 2: Define a feed/no feed

```{r Housekeeping - binding with light and ship data}
#Combine with light data 
#Load light data 
light <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Light/lightpollution_metric.csv")

#Night.of ID
sun <-getSunlightTimes(date = as.Date(light$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset"))
light <- cbind(light, sun)
light <- subset(light, select = -c(date, lat, lon))
light$night.of <- with(light, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

#Use average light pollution per night as metric 
light_night <- light %>%
  group_by(night.of) %>%
  summarise(
    avg_lightpoll = mean(lightpollution, na.rm = TRUE),
    med_lightpoll = median(lightpollution, na.rm = TRUE),
    moonPhase = median(moonPhase, na.rm = TRUE),
    med_illumination = median(illumination, na.rm = TRUE), 
    avg_illumination = mean(illumination, na.rm = TRUE),
    avg_moonlight = mean(moonlightModel, na.rm = TRUE),
    med_moonlight = median(moonlightModel, na.rm = TRUE),
    avg_logcdm2 = mean(logcdm2, na.rm = TRUE),
    medlogcdm2 = median(logcdm2, na.rm = TRUE)
  ) %>%
  ungroup()

#Create a night.of ID for chick weights too
meal_size <- meal_size %>%
  mutate(night.of = as.Date(POSIXct))

#Bind with chick meals 
meal_size <- meal_size %>%
  left_join(light_night, by = "night.of")

#Combine with wide ship data 
ship <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Light/Ship/Ship.csv")
#Night.of ID
sun <-getSunlightTimes(date = as.Date(ship$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset"))
ship <- cbind(ship, sun)
ship <- subset(ship, select = -c(date, lat, lon))
ship$night.of <- with(ship, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

#Average per night.of for heading, distance and number of ships
ship_night <- ship %>%
  group_by(night.of) %>%
  summarise(
    avg_heading = mean(heading_sk, na.rm = TRUE),
    avg_distance = mean(distance.metres, na.rm = TRUE),
    no_ships = max(Number.of.ships, na.rm = TRUE)) %>%
  ungroup()

#Bind with chick meals 
meal_size <- meal_size %>%
  left_join(ship_night, by = "night.of")
```

```{r Housekeeping - remove outliers}
#There is a relationship between meal size and hours
ggplot(data = data,
       aes(x = hours, y = meal_size))+
  geom_point(size = 1.0, alpha = 0.8)+
  geom_smooth(method = "lm", colour = "red") 
model <- lm(meal_size ~ hours, data = data)
summary(model)

#Exclude influential points that are creating the relationship 
#Exclude any with hours above 30 or below 16 
data_clean <- data[data$hours >= 19 & data$hours <= 30, ]
model_clean <- lm(meal_size ~ hours, data = data_clean)
summary(model_clean) #Relationship removed
ggplot(data = data_clean,
       aes(x = hours, y = meal_size))+
  geom_point(size = 1.0, alpha = 0.8)+
  geom_smooth(method = "lm", colour = "red") 
```

```{r Binary Feed}
#Add a column for fed and not fed and any meal size with a negative becomes 0 
#Any meal sizes that cross 0 in upper and lower bounds = no feed 

#Remove when hours between weighs is more than 40 (this is when the chick was OOR, we forgot etc - too long of a period for the equation to work) 

meal_size <- meal_size %>%
  mutate(
    feed = if_else(meal_lower > 0 & meal_upper > 0, "yes", "no")
  ) %>%
  filter(hours < 40)

#### Binary yes or no feed
#Plot to show which feeds were classified as 0 earlier (In)
ggplot(data_clean, aes(x = hours, y = meal_size, color = feed)) +
  geom_linerange(aes(ymin = meal_lower, ymax = meal_upper), alpha = 0.3, linewidth = 0.6) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40")+
  scale_colour_manual(
    name   = "Feed condition",
    values = c("no" = "red", "yes" = "grey50"),
    labels = c("Unfed", "Fed")) +
  scale_x_continuous(
    name = "Time between weighs (hours)",
    breaks = seq(0, max(data_clean$hours, na.rm = TRUE), by = 2)) +
  scale_y_continuous(
    name = "Meal size (g)",
    breaks = pretty(data_clean$meal_size)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 3))) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12))

#Save
write.csv(meal_size, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Chick Weights/mealsize.csv", row.names = FALSE)
```

#### Step 3: Fit a binomial regression

```{r Binomial Regression}
#Convert yes and no to 1 and 0
data_clean$feed_binary <- ifelse(data_clean$feed == "yes", 1, 0)

#Diagnostic Plot 
m <- lm(feed_binary ~ medlogcdm2, data=data_clean )
data_clean %>% mutate(yhat = predict(m)) %>%
  ggplot(aes(x=medlogcdm2, y=feed_binary)) + 
  geom_point() +
  geom_line(aes(y=yhat), color='red')

#Glm binomial model to define successes and failures
m1 <- glmer(cbind(feed_binary, 1-feed_binary) ~ medlogcdm2 + chick_age + (1|Burrow) + (1 | night.of), data=data_clean, family=binomial ) 
summary(m1)

#Get predictions 
emmeans(m1, ~medlogcdm2, at=list(medlogcdm2=1), type='response')

#Plot a graph
yhat.df <- emmeans(m1, ~medlogcdm2, at=list(medlogcdm2=seq(-3.572576, -1.794576,by=.01)), type='response') %>%
  as.data.frame()

ggplot(data_clean, aes(x=medlogcdm2)) +
  geom_ribbon(data=yhat.df, aes(ymin=asymp.LCL, ymax=asymp.UCL), fill='salmon', alpha=.4) +
  geom_line( data=yhat.df, aes(y=prob), color='red') +
  geom_point(aes(y=feed_binary)) +
  labs(y='Probability of Feed', x='Light levels (CDM2)')

#Plot of binomial regression
probcdm <- ggplot(data_clean, aes(x = medlogcdm2, y = feed_binary)) +
  stat_summary_bin(fun = mean, bins = 20, geom = "point", size = 1.5) +
  stat_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = TRUE,
              color = "blue",
              size = 1.2) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))"))+
  scale_y_continuous(
    name = "Probability of Feed") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12),
    legend.position = "none")
```

```{r Odds ratio}
#Odds Ratio = compares the likelihood of an outcome occuring in 2 groups 
#OR = 1 = no effect
#OR >1 = increased odds
#OR <1 = decreased odds

#Extract fixed effects estimate and SE for medlogcdm2
beta <- fixef(m1)["medlogcdm2"] #Before exponentiating (-0.42 negative so feeding odds decrease)
se   <- sqrt(vcov(m1)["medlogcdm2","medlogcdm2"]) #sqrt of variance = standard error 

# Odds ratio data
df_or <- data.frame(
  term  = "medlogcdm2",
  or    = exp(beta),
  lower = exp(beta - 1.96 * se),
  upper = exp(beta + 1.96 * se))

#Odds ratio plot
boxcdm <- ggplot(df_or, aes(x = term, y = or)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Odds ratio (95% CI)"
  ) +
  theme_minimal(base_size = 14)

#Patching figures together 
(boxcdm | probcdm) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```

## Hypothesis 8

Chicks are fed smaller meals on nights with brighter skies.

```{r Fixed Effect Creation}
#Days since start
data_clean$days_from_start <-as.numeric(as.Date(data_clean$night.of) - min(as.Date(data_clean$night.of), na.rm = TRUE))
```

```{r Averaging per night}
mean <- data_clean %>%
  group_by(night.of) %>%
  summarise(
    meal_size = mean(meal_size, na.rm = TRUE),
    mean_logcdm2 = mean(medlogcdm2, na.rm = TRUE),
    med_logcdm2 = median(medlogcdm2, na.rm = TRUE),
    med_illumination = median(med_illumination, na.rm = TRUE),
    moonPhase = mean(moonPhase, na.rm = TRUE),
    mean_moonlightModel = mean(med_moonlight, na.rm = TRUE),
    days_from_start = first(days_from_start)
  ) %>%
  ungroup()
```

```{r Model Fitting}
m2 <- lm(meal_size ~ med_logcdm2 + med_illumination + days_from_start, data = mean) 
summary(m2)

#Prediction framwork
newdata <- data.frame(
  med_logcdm2 = seq(min(mean$med_logcdm2), max(mean$med_logcdm2), length.out = 100),
  med_illumination = mean(mean$med_illumination, na.rm = TRUE),
  days_from_start = mean(mean$days_from_start, na.rm = TRUE))

pred_out <- predict(m2, newdata = newdata, se.fit = TRUE)

newdata$fit <- pred_out$fit
newdata$se  <- pred_out$se.fit
newdata$lwr <- newdata$fit - 1.96 * newdata$se
newdata$upr <- newdata$fit + 1.96 * newdata$se

#Plot 
ggplot(mean, aes(x = med_logcdm2, y = meal_size)) +
  geom_point() +
  geom_ribbon(
    data = newdata,
    aes(x = med_logcdm2, ymin = lwr, ymax = upr),
    inherit.aes = FALSE,
    alpha = 0.2) +
  geom_line(
    data = newdata,
    aes(x = med_logcdm2, y = fit),
    inherit.aes = FALSE,
    linewidth = 1, colour = "blue") +
  scale_y_continuous(
    name = "Mean Meal Size (g)",
    breaks = seq(0, max(mean$meal_size, na.rm = TRUE), by = 10)) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
```
