---
title: "Objective4"
output: html_document
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This file explores Objective 4 of the project: To evaluate whether shearwater behaviour varies within a night in response to short-term fluctuations in light levels. The file contains processing, analyses and plotting used for the final project write up.

## **Housekeeping** 

```{r Install and Load Packages}

#General
require(hms) #Time zone and handling times
require(dplyr) 
require(lubridate)
require(tidyverse)
require(purrr)
require(suncalc) #Calculates sun and moon elevation, rise and set
require(stringr)

#Plots
require(ggplot2) 
require(patchwork) #Compiling plots

#Models
require(MASS) #Predictions of linear models (glm.nb)
require(lme4) #Fit linear and generalised linear mixed models (glmer.nb)
require(ggfortify) #Plotting of statistics/inspecting residuals
require(car)

#Sound file processing 
require(tuneR)
require(seewave)
```

## Hypothesis 9

Shearwaters wait for the moon to set before arriving on the colony.

#### Step 1: Extract time of first call/time of first arrival

```{r Extract Time of First Arrival}
#Load data 
detections <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/Detections.csv")

#Need to first create a night column because my data points cross midnight
detections <- detections %>%
  mutate(
    #Combine date + time into a datetime
    datetime = ymd_hms(paste(UTC.Date, UTC.Time)),
    
    #Extract hour
    hour = hour(datetime),
    
    #Night begins at 22:00; hours < 12 considered the early morning of the same night
    Night = ifelse(hour >= 21, UTC.Date, UTC.Date-1))

#Convert night to date as it is numeric rn 
detections$Night <- as.Date(detections$Night)

#Extract the first detection per night
first_detection_per_night <- detections %>%
  arrange(Night, datetime) %>%
  group_by(Night) %>%
  filter(detection.no > 1) %>%
  slice_head(n = 1) %>%
  ungroup()

#Save to directory
write.csv(first_detection_per_night, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/first_detection_per_night.csv", row.names = FALSE)
```

#### Step 2: Test how well does BirdNET detect time of first call? 

```{r Time of First Call - Comparison of Manual vs Automated BirdNET detections}
############## Get a random  sample of BirdNET files for manual counts
#Read in the csv that has time of arrival w the rest of the data
arrival <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/first_detection_per_night.csv")

#Removd NAs (where we dont have time of arrival data)
samplearrival <- subset(arrival,!is.na(Time.of.first.arrival))

#Obtain random sample of 10
samplearrival <- samplearrival[sample(nrow(samplearrival), 10), ]

#Save as csv for counting manually (outside of R)
write.csv(samplearrival, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/samplearrival.csv", row.names = FALSE)

#Read in csv with my manual and automated time of arrival 
samplearrival <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/tests/manualarrival/samplearrival.csv")

#Order in which the time is in the file (hour:min:sec)
orders <- c("H:M:S")

#Function that safely processes the time depending on the format 
parse_time_safe <- function(x) {
  #Return POSIXct or NA
  #First try hms() (time-only), then parse_date_time with multiple orders
  out <- suppressWarnings(hms::as_hms(x))
  if (all(is.na(out))) {
    out2 <- suppressWarnings(parse_date_time(x, orders = orders, tz = "UTC"))
    # If parse_date_time returns date-times, return them
    if (all(is.na(out2))) return(NA)
    return(out2)
  } else {
    return(out)
  }
}

#Apply parse_date_time directly and keep POSIXct result:
parsed_manual <- suppressWarnings(parse_date_time(samplearrival$Manual.Time.of.Arrival, orders = orders, tz = "UTC"))
parsed_bird <- suppressWarnings(parse_date_time(samplearrival$Time.of.first.arrival, orders = orders, tz = "UTC"))

#If your values are time-of-day only, parse_date_time will parse them to today + time.
#Inspect parse success rate:
cat("Manual parsed NAs:", sum(is.na(parsed_manual)), " / ", length(parsed_manual), "\n")
cat("BirdNET parsed NAs:", sum(is.na(parsed_bird)), " / ", length(parsed_bird), "\n")

#Convert parsed values to minutes since midnight
to_minutes <- function(dt) {
  if (inherits(dt, "Period") || inherits(dt, "difftime")) {
    return(as.numeric(dt) / 60)
  }
  # If POSIXct, extract hour/minute/second
  h <- hour(dt)
  m <- minute(dt)
  s <- second(dt)
  # For NA dt this yields NA components and therefore NA total
  return(h*60 + m + s/60)
}

manual_sec <- to_minutes(parsed_manual)
bird_sec   <- to_minutes(parsed_bird)

#Combine rows
df <- data.frame(manual_sec = manual_sec, bird_sec = bird_sec)

#Plot
arrival <- if (nrow(nooutlier) > 0) {
  
  ggplot(nooutlier, aes(x = manual_sec, y = bird_sec)) +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, color = "red") +  # 1:1 line
    geom_smooth(method = "lm", se = FALSE, color = "black") +  # linear model
    labs(
      x = "Manual (minutes since midnight)",
      y = "BirdNET Model (minutes since midnight)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.title = element_text(size = 12),
      axis.text  = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.position = c(0,1),
      legend.justification = c(0, 1),
      legend.background = element_rect(fill = "white", color = "black"),
      legend.key = element_rect(fill = "white", color = NA)
    )
  
} else {
  stop("No valid (non-NA) parsed times found for both columns. Check formats and parsing.")
}

```

#### Step 3: Convert dates into moon cycles 

```{r}
#Split data into moon cyles (time of night when moon is at its highest and therefore brightest for each moon cycle)
#Get moon illumination info
moon <- getMoonIllumination(date = data$UTC.Date)

#PhaseNames by the phase
phase_name <- function(p) {
  if (p < 0.0625 || p >= 0.9375) return("New Moon")
  if (p < 0.1875) return("Waxing Crescent")
  if (p < 0.3125) return("First Quarter")
  if (p < 0.4375) return("Waxing Gibbous")
  if (p < 0.5625) return("Full Moon")
  if (p < 0.6875) return("Waning Gibbous")
  if (p < 0.8125) return("Last Quarter")
  return("Waning Crescent")
}
data$moonPhaseName <- vapply(moon$phase, phase_name, character(1))

#Get moon rise and set 
data$UTC.Date <- as.Date(data$UTC.Date, format = "%Y-%m-%d")
lat <- 51.737449
lon <- -5.287829
moon_times <- data %>%
  distinct(UTC.Date) %>%
  mutate(
    times = map(UTC.Date, ~ getMoonTimes(.x, lat = lat, lon = lon)),
    moonrise = map(times, "rise"),
    moonset  = map(times, "set")
  ) %>%
  select(-times)

data <- data %>%
  inner_join(moon_times, by = "UTC.Date")

data$moonset <- as.POSIXct(unlist(data$moonset), origin = "1970-01-01", tz = "UTC")

data$moonrise <- as.POSIXct(unlist(data$moonrise), origin = "1970-01-01", tz = "UTC")
```

#### Step 4: Fit a model 

```{r Aesthetics}
#Order labels so they plot in moon cycle order
data$moonPhaseName <- factor(
  data$moonPhaseName,
  levels = rev(c(
    "New Moon", "Waxing Crescent", "First Quarter",
    "Waxing Gibbous", "Full Moon",
    "Waning Gibbous", "Last Quarter",
    "Waning Crescent",
    "ALAN")))

#Colours
phase_cols <- c(
  "#8DA0CB",
  "#FFD92F",
  "#66C2A5",
  "#FC8D62",
  "#E78AC3",
  "#A6D854",
  "#E5C494",
  "#B3B3B3")

```

```{r Fixed Effect Creation}
###### Days from Start 
data$days_from_start <-as.numeric(as.Date(data$Night) - min(as.Date(data$Night), na.rm = TRUE))

###### Hourse Since Midnight
data$hour <- as.numeric(substr(data$UTC.Time, 1, 2)) +
  as.numeric(substr(data$UTC.Time, 4, 5)) / 60 +
  as.numeric(substr(data$UTC.Time, 7, 8)) / 3600
```

```{r Model Fitting}
#Model
firstcall <- glmmTMB(hour ~ moonPhaseName + days_from_start, family = Gamma(link = "log"), data = data)
summary(firstcall)

#Pairwise comparisons from the fitted model
emm <- emmeans(firstcall, ~ moonPhaseName)
pw <- as.data.frame(pairs(emm, adjust = "tukey"))

#Split "A - B" into two columns for plotting brackets
pw <- pw %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    p.signif = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      TRUE ~ "ns"))

#Set bracket heights
ymax <- max(data$hour, na.rm = TRUE)
yrng <- diff(range(data$hour, na.rm = TRUE))
pw2 <- pw %>%
  dplyr::filter(
    group1 %in% c("First Quarter", "Last Quarter"),
    group2 %in% c("First Quarter", "Last Quarter")
  ) %>%
  dplyr::mutate(y.position = 23.8)

#Calc average time of moon set and rise for each moon phase 
data$set_hour <- as.numeric(format(data$moonset, "%H")) +
  as.numeric(format(data$moonset, "%M")) / 60 +
  as.numeric(format(data$moonset, "%S")) / 3600
data$rise_hour <- as.numeric(format(data$moonrise, "%H")) +
  as.numeric(format(data$moonrise, "%M")) / 60 +
  as.numeric(format(data$moonrise, "%S")) / 3600

#Cyclical mean (taking into account crossing midnight)
data %>%
  group_by(moonPhaseName) %>%
  summarise(
    set_mean = {
      theta <- set_hour / 24 * 2 * pi
      mt <- atan2(
        mean(sin(theta), na.rm = TRUE),
        mean(cos(theta), na.rm = TRUE))
      (mt %% (2*pi)) / (2*pi) * 24},
    rise_mean = {
      theta <- rise_hour / 24 * 2 * pi
      mt <- atan2(
        mean(sin(theta), na.rm = TRUE),
        mean(cos(theta), na.rm = TRUE))
      (mt %% (2*pi)) / (2*pi) * 24})

#Use these in geom_segment in ggplot to mark the average moon rise an set times for the moon phases (number are manually taken from this output)
```

```{r Plot}
#only plotting Last and First Quarter phases 
ggplot(data, aes(x = moonPhaseName, y = hour, fill = moonPhaseName)) +
  geom_boxplot(outlier.shape = NA, width = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  scale_fill_manual(values = phase_cols) +
  #scale_color_manual(values = phase_cols) +
  scale_x_discrete(name = "Moon Phase", limits = c("First Quarter", "Last Quarter")) +
  scale_y_continuous(
    name = "Time of First Call",
    breaks = seq(21, 24, by = 1),
    labels = function(x) sprintf("%02d:00", x)) +
  coord_cartesian(ylim = c(21, 24)) +
  stat_pvalue_manual(
    pw2,
    label = "p.signif",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    hide.ns = TRUE,
    size = 7) +
  geom_segment(
    aes(x = 0.6, xend = 1.4, y = 23.1, yend = 23.1),
    linetype = "dashed",
    color = "#FFD92F",
    linewidth = 1,
    inherit.aes = FALSE) +
  geom_segment(
    aes(x = 1.6, xend = 2.4, y = 22.6, yend = 22.6),
    linetype = "dashed",
    color = "#8DA0CB",
    linewidth = 1,
    inherit.aes = FALSE) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_markdown(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 15, margin = margin(r=20)),
    axis.title.x = element_text(size = 15, margin = margin(t=20)))

```
