---
title: "Objective2"
output: html_document
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This file explores Objective 2 of the project: [To determine whether light and ships impact activity levels at the breeding colony]{.underline}. The file contains processing, analyses and plotting used for the final project write up.

## **Housekeeping**

```{r Install and Load packages}

#General
require(hms) #Time zone and handling times
require(dplyr) 
require(lubridate)
require(tidyverse)
require(purrr)
require(suncalc) #Calculates sun and moon elevation, rise and set
require(stringr)

#Plots
require(ggplot2) 
require(patchwork) #Compiling plots

#Models
require(MASS) #Predictions of linear models (glm.nb)
require(lme4) #Fit linear and generalised linear mixed models (glmer.nb)
require(ggfortify) #Plotting of statistics/inspecting residuals
require(car)

#Sound file processing 
require(tuneR)
require(seewave)
```

## Hypothesis 4

The number of shearwaters on the surface of the colony is significantly lower when sky brightness is higher

```{r Load Data}
#Hypothesis 4 (surface activity) data
scopecount <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Analysis/analysis/data/CountsSurface.csv")
head(scopecount)

#Midnight values for POSIXct are missing (due to file saving) so add back 
scopecount$POSIXct <- as.POSIXct(
  ifelse(nchar(as.character(scopecount$POSIXct)) == 10,
         paste(scopecount$POSIXct, "00:00:00"),
         as.character(scopecount$POSIXct)),
  format="%Y-%m-%d %H:%M:%S")
```

```{r Fixed and Random Effect Creation}
########## Days from Start
#Create a Night Of ID (because data crossed midnight)
#Coordinates of study site to get times of sunset and sunrise 
sun <-getSunlightTimes(date = as.Date(scopecount$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset")) 
scopecount <- cbind(scopecount, sun)
scopecount <- subset(scopecount, select = -c(date, lat, lon))

#Set night of as the time between sunset and sunrise the next day
scopecount$night.of <- with(scopecount, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

#Calculate days from start
scopecount$days_from_start <-as.numeric(as.Date(scopecount$night.of) - min(as.Date(scopecount$night.of), na.rm = TRUE))

########## Hours since Midnight
#Include time of day 
scopecount$hours_since_midnight <- hour(scopecount$POSIXct) + minute(scopecount$POSIXct)/60 + second(scopecount$POSIXct)/3600

head(scopecount)
```

```{r Model Testing}
#Count data - a Poisson distribution with log link function (makes sure predictions are not negative)

#First model using poisson distribution and log link 
glm1 <- glm(Number.on.surface ~ logcdm2 + days_from_start + hours_since_midnight, family = poisson(link = "log"), data = scopecount)
autoplot(glm1, which = c(1,3))
summary(glm1)
#Did not use this model:
  #Variance is not equal to the mean (ratio of residual deviance to residual degrees of         freedom) 
  #The ratio of residual deviance to df is 1161.8:246 which is really bad (1161.8:246 is much   larger than 1) 

#This large ratio indicates overdispersion
#Is the data zero-inflated? (common cause of overdispersion)
table(scopecount$Number.on.surface) #Yes it is so need to fit a (1) quasi poisson or (2) negative binomial

#Solution 1: Quasi poisson model 
glm2 <- glm(Number.on.surface ~ logcdm2 + days_from_start + hours_since_midnight, 
            family = quasipoisson(link = "log"), 
            data = scopecount)
autoplot(glm2, which = c(1,3))
summary(glm2)
#Still really high ratio - not good model 

#Solution 2 is a negative binomial 
glm3 <- glm.nb(Number.on.surface ~ logcdm2 + days_from_start + hours_since_midnight, 
               data = scopecount)
autoplot(glm3, which = c(2))
summary(glm3)
#Better!

#Add a random effect of night.of bc need to account for replication within nights  
#This is a generalized mixed effects model (GLMM) with negative binomial distribution and log link
glm4 <- glmer.nb(Number.on.surface ~ logcdm2 + illumination + days_from_start + hours_since_midnight + (1 | night.of),
         data = scopecount)
summary(glm4)
#This made the residual deviance 294.55:246 which is much better 

#AIC also shows glm4 is better than glm1 (glm1 = 2002, glm4 = 1422)
AIC(glm1, glm2, glm3, glm4)
```

```{r Model and Plot - Including Natural Illumination}
#Final model used: 
allglm <- glmer.nb(Number.on.surface ~ no.ships + logillumination.x + logcdm2.x + days_from_start + hours_since_midnight + (1 | night.of), data = scopecount)
summary(allglm)

#Prediction dataframe for cd/m2
cdm_all <- data.frame(
  logcdm2.x = seq(min(scopecount$logcdm2.x, na.rm = TRUE),
                  max(scopecount$logcdm2.x, na.rm = TRUE),
                  length.out = 100),
  no.ships = mean(scopecount$no.ships, na.rm = TRUE),
  logillumination.x = mean(scopecount$logillumination.x, na.rm = TRUE),
  days_from_start = mean(scopecount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(scopecount$hours_since_midnight, na.rm = TRUE))

#Prediction dataframe for natural illumination
nat_all <- data.frame(
  logillumination.x = seq(min(scopecount$logillumination.x, na.rm = TRUE),
                  max(scopecount$logillumination.x, na.rm = TRUE),
                  length.out = 100),
  no.ships = mean(scopecount$no.ships, na.rm = TRUE),
  logcdm2.x = mean(scopecount$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(scopecount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(scopecount$hours_since_midnight, na.rm = TRUE))

#Prediction data for ships
ship_all <- data.frame(
  no.ships = seq(min(scopecount$no.ships, na.rm = TRUE),
                          max(scopecount$no.ships, na.rm = TRUE),
                          length.out = 100),
  logillumination.x = mean(scopecount$logillumination.x, na.rm = TRUE),
  logcdm2.x = mean(scopecount$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(scopecount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(scopecount$hours_since_midnight, na.rm = TRUE))

#Predict for cd/m2
cdm_all$fit_link <- predict(allglm, cdm_all, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logillumination.x + logcdm2.x + days_from_start + hours_since_midnight, cdm_all)
V <- vcov(allglm)
cdm_all$se_link <- sqrt(diag(X %*% V %*% t(X)))
cdm_all$lwr <- exp(cdm_all$fit_link - 1.96 * cdm_all$se_link)
cdm_all$upr <- exp(cdm_all$fit_link + 1.96 * cdm_all$se_link)
cdm_all$pred <- exp(cdm_all$fit_link)

#Predict for natural illumination
nat_all$fit_link <- predict(allglm, nat_all, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logillumination.x + logcdm2.x + days_from_start + hours_since_midnight, nat_all)
V <- vcov(allglm)
nat_all$se_link <- sqrt(diag(X %*% V %*% t(X)))
nat_all$lwr <- exp(nat_all$fit_link - 1.96 * nat_all$se_link)
nat_all$upr <- exp(nat_all$fit_link + 1.96 * nat_all$se_link)
nat_all$pred <- exp(nat_all$fit_link)

#Predict for ships
ship_all$fit_link <- predict(allglm, ship_all, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logillumination.x + logcdm2.x + days_from_start + hours_since_midnight, ship_all)
V <- vcov(allglm)
ship_all$se_link <- sqrt(diag(X %*% V %*% t(X)))
ship_all$lwr <- exp(ship_all$fit_link - 1.96 * ship_all$se_link)
ship_all$upr <- exp(ship_all$fit_link + 1.96 * ship_all$se_link)
ship_all$pred <- exp(ship_all$fit_link)

#Plot ships
shipplot1 <- ggplot(scopecount, aes(x = no.ships, y = Number.on.surface)) +
  geom_point() +
  geom_line(data = ship_all, aes(x = no.ships, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = ship_all,
              aes(x = no.ships, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters on Surface",
    breaks = seq(0, max(scopecount$Number.on.surface, na.rm = TRUE), by = 10)) +
  coord_cartesian(ylim = c(0, 40)) +
  scale_x_continuous(
    name = expression("Number of Ships")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
shipplot1

#Plot cd/m2
cdmplot1 <- ggplot(scopecount, aes(x = logcdm2.x, y = Number.on.surface)) +
  geom_point() +
  geom_line(data = cdm_all, aes(x = logcdm2.x, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = cdm_all,
              aes(x = logcdm2.x, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters on Surface",
    breaks = seq(0, max(scopecount$Number.on.surface, na.rm = TRUE), by = 10)) +
  coord_cartesian(ylim = c(0, 40)) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
cdmplot1

#Plot natural illumination
natplot <- ggplot(scopecount, aes(x = logillumination.x, y = Number.on.surface)) +
  geom_point() +
  geom_line(data = nat_all, aes(x = logillumination.x, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = nat_all,
              aes(x = logillumination.x, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters on Surface",
    breaks = seq(0, max(scopecount$Number.on.surface, na.rm = TRUE), by = 10)) +
  coord_cartesian(ylim = c(0, 40)) +
  scale_x_continuous(
    name = expression("Natural Illumination (log(lux))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
natplot

```

```{r Model and Plot - Excluding Natural Illumination}
#Filter to remove when natural illuminance is above 0
filtered <- scopecount %>%
  filter(illumination.x <= 0)

#Model used without natural illumination as a factor 
filteredglm <- glmer.nb(Number.on.surface ~ no.ships + logcdm2.x + days_from_start + hours_since_midnight + (1 | night.of), data = filtered)
summary(filteredglm)
Anova(filteredglm, type = "II")

#Create prediction data for cd/m2
cdm_new <- data.frame(
  logcdm2.x = seq(min(filtered$logcdm2.x, na.rm = TRUE),
                  max(filtered$logcdm2.x, na.rm = TRUE),
                  length.out = 100),
  no.ships = mean(filtered$no.ships, na.rm = TRUE),
  days_from_start = mean(filtered$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(filtered$hours_since_midnight, na.rm = TRUE))

#Create prediction data for ships
ship_new <- data.frame(
  no.ships = seq(min(filtered$no.ships, na.rm = TRUE),
                 max(filtered$no.ships, na.rm = TRUE),
                 length.out = 100),
  logcdm2.x = mean(filtered$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(filtered$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(filtered$hours_since_midnight, na.rm = TRUE))

#Predict for cd/m2
cdm_new$fit_link <- predict(filteredglm, cdm_new, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logcdm2.x + days_from_start + hours_since_midnight, cdm_new)
V <- vcov(filteredglm)
cdm_new$se_link <- sqrt(diag(X %*% V %*% t(X)))
cdm_new$lwr <- exp(cdm_new$fit_link - 1.96 * cdm_new$se_link)
cdm_new$upr <- exp(cdm_new$fit_link + 1.96 * cdm_new$se_link)
cdm_new$pred <- exp(cdm_new$fit_link)

#Predict for ships
ship_new$fit_link <- predict(filteredglm, ship_new, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logcdm2.x + days_from_start + hours_since_midnight, ship_new)
V <- vcov(filteredglm)
ship_new$se_link <- sqrt(diag(X %*% V %*% t(X)))
ship_new$lwr <- exp(ship_new$fit_link - 1.96 * ship_new$se_link)
ship_new$upr <- exp(ship_new$fit_link + 1.96 * ship_new$se_link)
ship_new$pred <- exp(ship_new$fit_link)

#Cd/m2 plot
cdmplot2 <- ggplot(filtered, aes(x = logcdm2.x, y = Number.on.surface)) +
  geom_point() +
  geom_line(data = cdm_new, aes(x = logcdm2.x, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = cdm_new,
              aes(x = logcdm2.x, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters on Surface",
    breaks = seq(0, max(filtered$Number.on.surface, na.rm = TRUE), by = 10))+
  coord_cartesian(ylim = c(0, 30)) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
cdmplot2

#Ship plot
shipplot2 <- ggplot(filtered, aes(x = no.ships, y = Number.on.surface)) +
  geom_point() +
  geom_line(data = ship_new, aes(x = no.ships, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = ship_new,
              aes(x = no.ships, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwaters on Surface",
    breaks = seq(0, max(filtered$Number.on.surface, na.rm = TRUE), by = 10))+
  coord_cartesian(ylim = c(0, 30)) +
  scale_x_continuous(
    name = expression("Number of Ships")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
shipplot2
```

```{r Patch together the figures}
(shipplot1 | cdmplot1) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")

(shipplot2 | cdmplot2) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```

## Hypothesis 5

Vocal activity is significantly lower when sky brightness is higher.

#### Step 1: High Pass Filtering

Removing any background noise from the sound files (mainly wind noise).

```{r High Pass Filtering}
#High pass filtering for bird frequencies (2-8kHz)

#Set working directory
setwd("/Users/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files")

#Base directories (raw_files contains my original sound files and filtered-files is where I want the new files to go that are filtered)
input_base <- "raw-files"
output_base <- "filtered-files"

#Find all .wav files within all subfolders. Case is important to distinguish because some files are .wav and some are .WAV
raw_wav_files <- list.files("raw-files", pattern = "\\.wav$", full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

#Check that the files are all being called with length(raw_wav_files). It calls around 3000 files so that sounds about right for the data set I have 
for (wav_path in raw_wav_files) {

#Extract relative path (e.g., "2025-07-01/Peak activity/file.wav") but removing the last path separator = .wav files (this is because we only want the folders not the files to be replicated in the output base)
  rel_path <- dirname(sub(paste0("^", input_base, "/"), "", wav_path))
  
  #Make a matching output folder structure (dir.create makes a file path)
  output_folder <- file.path(output_base, rel_path)
  dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)
  
  #Read the file
  file <- readWave(wav_path)
  
  #Apply bandpass filter
  filtered <- bwfilter(file, f = file@samp.rate, from = 230)
  
  #Convert to Wave object
  filtered_wave <- Wave(left = filtered, samp.rate = file@samp.rate, bit = file@bit)
  
  #Build output filename - basename removes all the whole whole path except the file name (.wav) and then we add filtered onto the front to distinguish it from the raw files
  file_name <- tools::file_path_sans_ext(basename(wav_path))
  output_file <- file.path(output_folder, paste0(file_name, "-filtered.wav"))
  
  #Save filtered file
  savewav(filtered_wave, filename = output_file)
  #Print what files are saved so I can keep track of progress as it downloads - e.g. needed to troubleshoot as stopped in the middle of the file
  cat("Saved filtered file to:", output_file, "\n")}
```

#### Step 2: Which BirdNET model is best?

```{r Comparing Models from BirdNET Analyser}
#Load all the csv files of the models (first 1 are slightltly different so require a diff approach)
modelpaths <- c(
  "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/pattern-detection/model-output-1/CustomClassifier_evaluation.csv",
  "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/pattern-detection/model-output-2/Model2_evaluation.csv",
  paste0("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/pattern-detection/model-output-", 3:10, "/CustomClassifier_evaluation.csv")
) |> unlist()

#Add an identifier for each model (e.g The first model = Model 1) 
#Merge them into one dataframe
all_models <- map2_dfr(modelpaths, seq_along(modelpaths), ~ {
  read.csv(.x) |> mutate(Model = paste0("Model", .y))})

#Extract model with highest f.score for identification of Manx shearwater
best_manx_model <- all_models %>%
  #Only showing scores for detecting Manx
  filter(Class == "Puffinus puffinus_Manx Shearwater") %>% 
  #Removig columns that arent relevant to analysis
  select(-Precision..0.5.,
         -Recall..0.5.,
         -F1.Score..0.5.,
         -Class) %>% 
  #Renaming columns
  rename(Precision = Precision..opt., 
         Recall = Recall..opt., 
         F1.Score = F1.Score..opt., 
         'Optimal Threshold' = Optimal.Threshold, 
         'True Positives' = True.Positives, 
         'True Negatives' = True.Negatives, 
         'False Positives' = False.Positives,
         'False Negatives' = False.Negatives,
         Percentage = Percentage....) %>% 
  #Reording columns
  select(Model, Precision, everything()) %>% 
  #Order by F1 Score
  arrange(desc(F1.Score))

###MODEL 10 IS THE BEST - highest f score and very high AUROC
#Save this  
write.csv(best_manx_model, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/pattern-detection/BestModels.csv", row.names = FALSE)
```

#### Step 3: Extract Number of Calls from BirdNET

```{r Extract Number of Calls from Best Model}
#BirdNET provides an output csv which lists the detections in each file 
#Need to extract the number of calls per file, to get the number of calls per time point 

path <- "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/BirdNET_output"

#Get all CSV file names
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)

#Read them all into a list of data frames
df_list <- setNames(lapply(files, read.csv),
                    basename(tools::file_path_sans_ext(files)))

#To determine number of calls, we need to count the number of rows in each dataframe and make that into a new data-frame
detections <- data.frame(
  file = names(df_list),
  detection.no = sapply(df_list, nrow))

#Extracting date and time from the file names 
detections <- detections %>%
  mutate(UTC.Date = str_sub(file, 1, 8),
    UTC.Time = str_sub(file, 10, 15)) %>%
  mutate(UTC.Date = as.Date(UTC.Date, format = "%Y%m%d"),
  UTC.Time = format(strptime(UTC.Time, format = "%H%M%S"), "%H:%M:%S"))

#Merge with light data
light <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/LightMeter.csv")

#Getting date and time in the same format as sound data
light$UTC.Date <- format(as.Date(light$UTC.Date, format = "%d/%m/%Y"), "%Y-%m-%d")
light$UTC.Time <- format(strptime(light$UTC.Time, format = "%I:%M %p"), "%H:%M:%S")

#Merging 
detections.light <- merge(detections, light, by=c("UTC.Date","UTC.Time"), all = TRUE)

#Removing rows with NAs 
detections.clean <- na.omit(detections.light)

#Keep first of duplicates (duplicates bc have 2 audiomoths deployed - sometimes the 2 audiomoths overlapping in records - they were consistent with one another ) 
detections.clean <- detections.clean %>%
  distinct(UTC.Date, UTC.Time, file, .keep_all = TRUE)

#Save this in directory for safe keeping
write.csv(detections.clean, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/Detections.csv", row.names = FALSE)
```

#### Step 4: How well does BirdNET detect the number of calls?

```{r Number of Detections - Comparison of Manual vs Automated BirdNET detections}
############## Get a random stratified sample of BirdNET files for manual counts 
#Ensure required column exists:
detections.clean$file
#Find out what minimum and maximum number of detections are before determining bins for stratifications
min(detections.clean$detection.no) #0
max(detections.clean$detection.no) #118

#Creating bins based on the number of detections
detections.bins <- detections.clean %>%
  mutate(detection.bin = cut(detection.no,
                        breaks = c(0, 20, 40, 60, 80, 100, 120),
                        include.lowest = TRUE,
                        labels = c("0-20", "20-40", "40-60", "60-80", "80-100", "100-120")))

#Proportional sampling depending on number in each bin
#Saying I need a total of 50 samples and each bin must have at least 1 sample in it
#Max shouldnt have more than is available in the bin 
samples_per_bin <- detections.bins %>%
  group_by(detection.bin) %>%
  summarise(stratum_n = n(), .groups = "drop") %>%
  mutate(
    samples_needed = pmax(1, round(stratum_n / total_n * 50)),  # min 1
    samples_needed = pmin(samples_needed, stratum_n))

#Drawing the sample 
sampled <- detections.bins %>%
  left_join(samples_per_bin, by = "detection.bin") %>%
  group_by(detection.bin) %>%
  sample_n(size = first(8.3), replace = FALSE) %>%
  ungroup()

#Save the samples bc when you run the code again, the samples will change every time
write.csv(sampled, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/BirdNET_output/sampled_files.csv", row.names = FALSE)

#################### Calculating difference between manual and automatic counts
#Done manual counts without code 

#Insert the new dataframe now that I have manually counted the number of calls 
sampled <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/tests/manualvsautomated/sampled_files.csv")

#Ensure numeric columns
#detection.no =  BirdNET
#manual.detection.no = manual 
sampled$detection.no <- as.numeric(sampled$detection.no)
sampled$manual.detection.no <- as.numeric(sampled$manual.detection.no)

#Differences
sampled <- sampled %>%
  mutate(
    diff = detection.no - manual.detection.no,
    absdiff = abs(detection.no - manual.detection.no),
    sqdiff = diff^2
  )

#Mean Absolute Error
MAE <- mean(sampled$absdiff)

#Mean Error (bias)
bias <- mean(sampled$diff)

#Root Mean Square Error
RMSE <- sqrt(mean(sampled$sqdiff))

#Pearson correlation (linear relationship)
pearson_r <- cor(sampled$detection.no, sampled$manual.detection.no, method = "pearson")

#Spearman correlation (rank-based, robust to outliers)
spearman_r <- cor(sampled$detection.no, sampled$manual.detection.no, method = "spearman")

#Normality test for differences
shap <- shapiro.test(sampled$diff)

#Choose paired test (t-test if normal, Wilcoxon if not)
if (shap$p.value > 0.05) {
  paired_test <- t.test(sampled$detection.no, sampled$manual.detection.no, paired = TRUE)
  test_used <- "Paired t-test"
} else {
  paired_test <- wilcox.test(sampled$detection.no, sampled$manual.detection.no, paired = TRUE)
  test_used <- "Wilcoxon signed-rank test"
}

#Summary of all these metrics
results <- list(
  n = nrow(sampled),
  MAE = MAE,
  Bias = bias,
  RMSE = RMSE,
  Pearson_r = pearson_r,
  Spearman_r = spearman_r,
  Normality_p_value = shap$p.value,
  Test_used = test_used,
  Test_p_value = paired_test$p.value)

results

#All of this means there is a strong positive relationship between BirdNET counts and my manual counts
#It overcounts slightly but that is fine - bc it is relative 
#$n = 48
#MAE = 21.04167
#Bias = 12.54167 - Means the BirdNET systematically overcounts (positives)
#RMSE = 26.97375 
#Pearson_r = 0.7519756
#Spearman_r = 0.7363127 #Strong positive correlation between the manual and automatic counts

#Plot to show they are highly correlated
ggplot(sampled, aes(x = manual.detection.no, y = detection.no)) +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid", color = "red") +
  scale_x_continuous(
    name = "Manual Count",
    breaks = seq(0, max(sampled$manual.detection.no, na.rm = TRUE), by = 20)
  ) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  scale_y_continuous(
    name = "BirdNET Model Count",
    breaks = pretty(sampled$detection.no)
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12))

#Which files are the outliers and why? 
#Biggest deviations from where manual = BirdNET detection (distance between manual and BirdNET detection)
sampled <- sampled %>%
  mutate(
    diff_identity = detection.no - manual.detection.no,
    abs_identity_dist = abs(diff_identity)
  )

#Worst differences (top 5)
top_outliers <- sampled %>%
  arrange(desc(abs_identity_dist)) %>%
  slice_head(n = 10)
top_outliers %>%
  select(file, manual.detection.no, detection.no, diff_identity)

#Highlighting these on the plot
plot(sampled$manual.detection.no, sampled$detection.no,
     xlab = "Manual count",
     ylab = "BirdNET count",
     main = "Manual vs BirdNET detections")
abline(lm(detection.no ~ manual.detection.no, data = sampled)) #Linear model
abline(0, 1, col = "red", lwd = 2) #If the relationship was perfectly correlated 
points(top_outliers$manual.detection.no,
       top_outliers$detection.no,
       col = "blue", pch = 19)
legend("topleft",
       legend = c("All files", "Largest deviations"),
       pch = c(1, 19),
       col = c("black", "blue"))

#Final plot used in dissertation using ggplot 
count <- ggplot(sampled, aes(x = manual.detection.no, y = detection.no)) +
  geom_point(aes(color = "All files"), size = 2) +
  geom_point(data = top_outliers,
             aes(x = manual.detection.no, y = detection.no, color = "Largest deviations"),
             size = 2) +
  geom_smooth(method = "lm", colour = "black", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_continuous(
    name = "Manual Count",
    breaks = seq(0, max(sampled$manual.detection.no, na.rm = TRUE), by = 20)
  ) +
  scale_y_continuous(
    name = "BirdNET Model Count",
    breaks = pretty(sampled$detection.no)
  ) +
  scale_color_manual(
    name = NULL,
    values = c("All files" = "black", "Largest deviations" = "blue")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = c(0,1),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA))

```

#### Step 5: Extract time of first call/time of first arrival

```{r Extract Time of First Arrival}
#Need to first create a night column because my data points cross midnight
detections <- detections %>%
  mutate(
    #Combine date + time into a datetime
    datetime = ymd_hms(paste(UTC.Date, UTC.Time)),
    
    #Extract hour
    hour = hour(datetime),
    
    #Night begins at 22:00; hours < 12 considered the early morning of the same night
    Night = ifelse(hour >= 21, UTC.Date, UTC.Date-1))

#Convert night to date as it is numeric rn 
detections$Night <- as.Date(detections$Night)

#Extract the first detection per night
first_detection_per_night <- detections %>%
  arrange(Night, datetime) %>%
  group_by(Night) %>%
  filter(detection.no > 1) %>%
  slice_head(n = 1) %>%
  ungroup()

#Save to directory
write.csv(first_detection_per_night, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/first_detection_per_night.csv", row.names = FALSE)
```

#### Step 6: How well does BirdNET detect time of first call?

```{r Time of First Call - Comparison of Manual vs Automated BirdNET detections}
############## Get a random  sample of BirdNET files for manual counts
#Read in the csv that has time of arrival w the rest of the data
arrival<- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/first_detection_per_night.csv")

#Removd NAs (where we dont have time of arrival data)
samplearrival <- subset(arrival,!is.na(Time.of.first.arrival))

#Obtain random sample of 10
samplearrival <- samplearrival[sample(nrow(samplearrival), 10), ]

#Save as csv for counting manually (outside of R)
write.csv(samplearrival, "~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/samplearrival.csv", row.names = FALSE)

#Read in csv with my manual and automated time of arrival 
samplearrival <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Data/Audiomoth files/tests/manualarrival/samplearrival.csv")

#Order in which the time is in the file (hour:min:sec)
orders <- c("H:M:S")

#Function that safely processes the time depending on the format 
parse_time_safe <- function(x) {
  #Return POSIXct or NA
  #First try hms() (time-only), then parse_date_time with multiple orders
  out <- suppressWarnings(hms::as_hms(x))
  if (all(is.na(out))) {
    out2 <- suppressWarnings(parse_date_time(x, orders = orders, tz = "UTC"))
    # If parse_date_time returns date-times, return them
    if (all(is.na(out2))) return(NA)
    return(out2)
  } else {
    return(out)
  }
}

#Apply parse_date_time directly and keep POSIXct result:
parsed_manual <- suppressWarnings(parse_date_time(samplearrival$Manual.Time.of.Arrival, orders = orders, tz = "UTC"))
parsed_bird <- suppressWarnings(parse_date_time(samplearrival$Time.of.first.arrival, orders = orders, tz = "UTC"))

#If your values are time-of-day only, parse_date_time will parse them to today + time.
#Inspect parse success rate:
cat("Manual parsed NAs:", sum(is.na(parsed_manual)), " / ", length(parsed_manual), "\n")
cat("BirdNET parsed NAs:", sum(is.na(parsed_bird)), " / ", length(parsed_bird), "\n")

#Convert parsed values to minutes since midnight
to_minutes <- function(dt) {
  if (inherits(dt, "Period") || inherits(dt, "difftime")) {
    return(as.numeric(dt) / 60)
  }
  # If POSIXct, extract hour/minute/second
  h <- hour(dt)
  m <- minute(dt)
  s <- second(dt)
  # For NA dt this yields NA components and therefore NA total
  return(h*60 + m + s/60)
}

manual_sec <- to_minutes(parsed_manual)
bird_sec   <- to_minutes(parsed_bird)

#Combine rows
df <- data.frame(manual_sec = manual_sec, bird_sec = bird_sec)

#Plot
arrival <- if (nrow(nooutlier) > 0) {
  
  ggplot(nooutlier, aes(x = manual_sec, y = bird_sec)) +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, color = "red") +  # 1:1 line
    geom_smooth(method = "lm", se = FALSE, color = "black") +  # linear model
    labs(
      x = "Manual (minutes since midnight)",
      y = "BirdNET Model (minutes since midnight)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.title = element_text(size = 12),
      axis.text  = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.position = c(0,1),
      legend.justification = c(0, 1),
      legend.background = element_rect(fill = "white", color = "black"),
      legend.key = element_rect(fill = "white", color = NA)
    )
  
} else {
  stop("No valid (non-NA) parsed times found for both columns. Check formats and parsing.")
}

```

#### Step 7: Model vocal activity with light/ship data

```{r Housekeeping}
#Load data
vocalcount <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Analysis/analysis/data/VocalDetections.csv")
  
#Adding back midnight time stamps
vocalcount$POSIXct <- as.POSIXct(
    ifelse(nchar(as.character(vocalcount$POSIXct)) == 10,
           paste(vocalcount$POSIXct, "00:00:00"),
           as.character(vocalcount$POSIXct)),
    format="%Y-%m-%d %H:%M:%S")

#Combine with ship data ship 
ship <- read.csv("data/shipwide.csv")

#Adding back midnight time stamps and POSIXct
ship$POSIXct <- as.POSIXct(
  ifelse(nchar(as.character(ship$POSIXct)) == 10,
         paste(ship$POSIXct, "00:00:00"),
         as.character(ship$POSIXct)),
  format="%Y-%m-%d %H:%M:%S")

#Bind with vocal detections
vocalcount <- vocalcount %>%
  inner_join(ship, by = "POSIXct")

colnames(vocalcount)
```

```{r Random and Fixed Effects Creation }
############ Days From Start 
#Night of ID already created above 
vocalcount$days_from_start <-as.numeric(as.Date(vocalcount$night.of) - min(as.Date(vocalcount$night.of), na.rm = TRUE))

############ Hours Since Midnight 
vocalcount$hours_since_midnight <- hour(vocalcount$POSIXct) + minute(vocalcount$POSIXct)/60 + second(vocalcount$POSIXct)/3600
```

```{r Model Fitting - Including Natural Illumination}
#Model - zero inflated data, negative binomial with log link 
glmvocal <- glmmTMB(detection.no ~ no.ships + logcdm2.x + logillumination.x + days_from_start + hours_since_midnight + (1 | night.of),
  ziformula = ~1,
  family = nbinom2,
  data = vocalcount)
summary(glmvocal)

#Prediction data for cdm
cdm_all <- data.frame(
  logcdm2.x = seq(min(vocalcount$logcdm2.x, na.rm = TRUE),
                  max(vocalcount$logcdm2.x, na.rm = TRUE),
                  length.out = 100),
  no.ships = mean(vocalcount$no.ships, na.rm = TRUE),
  logillumination.x = mean(vocalcount$logillumination.x, na.rm = TRUE),
  days_from_start = mean(vocalcount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(vocalcount$hours_since_midnight, na.rm = TRUE))

#Prediction data for illumination
nat_all <- data.frame(
  logillumination.x = seq(min(vocalcount$logillumination.x, na.rm = TRUE),
                          max(vocalcount$logillumination.x, na.rm = TRUE),
                          length.out = 100),
  no.ships = mean(vocalcount$no.ships, na.rm = TRUE),
  logcdm2.x = mean(vocalcount$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(vocalcount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(vocalcount$hours_since_midnight, na.rm = TRUE))

#Prediction data for ships
ship_all <- data.frame(
  no.ships = seq(min(vocalcount$no.ships, na.rm = TRUE),
                 max(vocalcount$no.ships, na.rm = TRUE),
                 length.out = 100),
  logillumination.x = mean(vocalcount$logillumination.x, na.rm = TRUE),
  logcdm2.x = mean(vocalcount$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(vocalcount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(vocalcount$hours_since_midnight, na.rm = TRUE))

#Predict Number of Ships
predships <- ggpredict(glmvocal, terms = "no.ships")
plot(predships)
shipplot1 <- ggplot() +
  geom_point(data = vocalcount,
             aes(x = no.ships, y = detection.no),
             size = 1.5) +
  geom_line(data = predships,
            aes(x = x, y = predicted),
            color = "blue", linewidth = 1) +
  geom_ribbon(data = predships,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  scale_x_continuous(
    name = "Number of Ships",
    breaks = seq(0, max(vocalcount$no.ships, na.rm = TRUE))) +
  scale_y_continuous(
    name = expression("Number of Shearwater Calls")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 12), 
    axis.title.x = element_text(margin = margin(t = 14)), 
    axis.title.y = element_text(margin = margin(r = 14)))

#Prediction grid for cdm2
cdm_all <- data.frame(
  logcdm2.x = seq(min(vocalcount$logcdm2.x, na.rm = TRUE),
                  max(vocalcount$logcdm2.x, na.rm = TRUE),
                  length.out = 200),
  no.ships = mean(vocalcount$no.ships, na.rm = TRUE),
  logillumination.x = mean(vocalcount$logillumination.x, na.rm = TRUE),
  days_from_start = mean(vocalcount$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(vocalcount$hours_since_midnight, na.rm = TRUE),
  night.of = vocalcount$night.of[1])
pr <- predict(glmvocal, newdata = cdm_all, type = "response", re.form = NA, se.fit = TRUE)
cdm_all$predicted <- pr$fit
cdm_all$conf.low  <- pr$fit - 1.96 * pr$se.fit
cdm_all$conf.high <- pr$fit + 1.96 * pr$se.fit

#Plot cd/m2
cdmplot1 <- ggplot() +
  geom_point(data = vocalcount,
             aes(x = logcdm2.x, y = detection.no),
             size = 1.5) +
  geom_line(data = cdm_all,
            aes(x = logcdm2.x , y = predicted),
            color = "blue", linewidth = 1) +
  geom_ribbon(data = cdm_all,
              aes(x = logcdm2.x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  scale_x_continuous(
    name = "Logcdm2",
    breaks = seq(0, max(vocalcount$logcdm2.x, na.rm = TRUE))) +
  scale_y_continuous(
    name = expression("Number of Shearwater Calls")) +
  coord_cartesian(ylim = c(0,120))+
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 12), 
    axis.title.x = element_text(margin = margin(t = 14)), 
    axis.title.y = element_text(margin = margin(r = 14)))
cdmplot1

```

```{r Model Fitting - Excluding Natural Illumination}
#Filtering data
#Filter to remove when natural illuminance is above 0
filtered <- vocalcount %>%
  filter(illumination.x <= 0)

#Model fit
filvocalglm <- glmer.nb(detection.no ~ no.ships + logcdm2.x + days_from_start + hours_since_midnight + (1 | night.of), data = filtered)
summary(filvocalglm)
Anova(filvocalglm, type = "II")

#Create prediction data for cdm
cdm_new <- data.frame(
  logcdm2.x = seq(min(filtered$logcdm2.x, na.rm = TRUE),
                  max(filtered$logcdm2.x, na.rm = TRUE),
                  length.out = 100),
  no.ships = mean(filtered$no.ships, na.rm = TRUE),
  days_from_start = mean(filtered$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(filtered$hours_since_midnight, na.rm = TRUE))

#Create prediction data for ship
ship_new <- data.frame(
  no.ships = seq(min(filtered$no.ships, na.rm = TRUE),
                 max(filtered$no.ships, na.rm = TRUE),
                 length.out = 100),
  logcdm2.x = mean(filtered$logcdm2.x, na.rm = TRUE),
  days_from_start = mean(filtered$days_from_start, na.rm = TRUE),
  hours_since_midnight = mean(filtered$hours_since_midnight, na.rm = TRUE))

#Predict for cdm
cdm_new$fit_link <- predict(filvocalglm, cdm_new, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logcdm2.x + days_from_start + hours_since_midnight, cdm_new)
V <- vcov(filvocalglm)
cdm_new$se_link <- sqrt(diag(X %*% V %*% t(X)))
cdm_new$lwr <- exp(cdm_new$fit_link - 1.96 * cdm_new$se_link)
cdm_new$upr <- exp(cdm_new$fit_link + 1.96 * cdm_new$se_link)
cdm_new$pred <- exp(cdm_new$fit_link)

#Predict for ship
ship_new$fit_link <- predict(filvocalglm, ship_new, type = "link", re.form = NA)
X <- model.matrix(~ no.ships + logcdm2.x + days_from_start + hours_since_midnight, ship_new)
V <- vcov(filvocalglm)
ship_new$se_link <- sqrt(diag(X %*% V %*% t(X)))
ship_new$lwr <- exp(ship_new$fit_link - 1.96 * ship_new$se_link)
ship_new$upr <- exp(ship_new$fit_link + 1.96 * ship_new$se_link)
ship_new$pred <- exp(ship_new$fit_link)

#Plot cd/m2
cdmplot2 <- ggplot(filtered, aes(x = logcdm2.x, y = detection.no)) +
  geom_point() +
  geom_line(data = cdm_new, aes(x = logcdm2.x, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = cdm_new,
              aes(x = logcdm2.x, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwater Calls",
    breaks = seq(0, max(filtered$detection.no, na.rm = TRUE), by = 10))+
  coord_cartesian(ylim = c(0, 30)) +
  scale_x_continuous(
    name = expression("Sky Brightness (log(" * cd/m^2 * "))")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
cdmplot2

#Plot ship 
shipplot2 <- ggplot(filtered, aes(x = no.ships, y = detection.no)) +
  geom_point() +
  geom_line(data = ship_new, aes(x = no.ships, y = pred),
            linewidth = 1, colour = "blue") +
  geom_ribbon(data = ship_new,
              aes(x = no.ships, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.2) +
  scale_y_continuous(
    name = "Number of Shearwater Calls",
    breaks = seq(0, max(filtered$detection.no, na.rm = TRUE), by = 10))+
  coord_cartesian(ylim = c(0, 30)) +
  scale_x_continuous(
    name = expression("Number of Ships")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 13), 
    axis.title.x = element_text(margin = margin(t = 12)), 
    axis.title.y = element_text(margin = margin(r = 12)))
shipplot2
```
