---
title: "Objective1"
output: html_document
date: "2026-04-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**

This file explores Objective 1 of the project: [To evaluate the impact of ships on sky brightness]{.underline}. The file contains processing, analyses and plotting used for the final project write up.

## **Housekeeping**

```{r Install and Load packages}
#General
require(hms) #Time zone and handling times
require(dplyr) 
require(lubridate)
require(tidyverse)
require(suncalc) #Calculates sun and moon elevation, rise and set
require(geosphere) #Calculates distance between 2 coordinates

#Plots
require(ggplot2) 
require(ggtext) #Text labels
require(patchwork) #Compiling plots

#Models
require(glmmTMB)
require(broom.mixed) #Residuals and confidence intervals
require(emmeans) #Pairwise comparisons between groups 
require(ggeffects) #Getting predictions and plotting them
```

```{r Load Data}
data <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Analysis/analysis/data/Ship.csv")
head(data)

#Midnight values for POSIXct are missing (due to file saving) so add back 
data$POSIXct <- as.POSIXct(
  ifelse(nchar(as.character(data$POSIXct)) == 10,
         paste(data$POSIXct, "00:00:00"),
         as.character(data$POSIXct)),
  format="%Y-%m-%d %H:%M:%S")
```

```{r Fixed and Random Effect Creation}

########## Days from Start
#Create a Night Of ID (because data crossed midnight)
#Coordinates of study site to get times of sunset and sunrise 
sun <-getSunlightTimes(date = as.Date(data$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset")) 
data <- cbind(data, sun)
data <- subset(data, select = -c(date, lat, lon))

#Set night of as the time between sunset and sunrise the next day
data$night.of <- with(data, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

#Calculate days from start
data$days_from_start <-as.numeric(as.Date(data$night.of) - min(as.Date(data$night.of), na.rm = TRUE))

########## Hours since Midnight 
data$hours_since_midnight <- hour(data$POSIXct) + minute(data$POSIXct)/60 + second(data$POSIXct)/3600

head(data)

########## Filter to remove natural illumination 
darkdata <- filter(data, illumination == 0)

```

## **Hypothesis 1**

**Sky brightness increases with the number of ships present in the area.**

```{r Model Creation and Output}
#Model
shipmodel <- glmmTMB(
  cdm2 ~ Number.of.ships + days_from_start + hours_since_midnight + (1 | days_from_start), 
  family = Gamma(link = "log"), 
  data = darkdata)

#Model output
summary(shipmodel)
pred <- ggpredict(shipmodel, terms = "Number.of.ships")
plot(pred)

#Plot the model
ship <- ggplot() +
  geom_point(data = darkdata,
             aes(x = Number.of.ships, y = cdm2),
             size = 1.5) +
  geom_line(data = pred,
            aes(x = x, y = predicted),
            color = "blue", linewidth = 1) +
  geom_ribbon(data = pred,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  scale_x_continuous(
    name = "Number of Ships",
    breaks = seq(0, max(darkdata$Number.of.ships, na.rm = TRUE), by = 1)) +
  scale_y_continuous(
    name = expression("Sky Brightness (" * cd/m^2 * ")")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 14), 
    axis.title.x = element_text(margin = margin(t = 14)), 
    axis.title.y = element_text(margin = margin(r = 14)))
```

## **Hypothesis 2**

**Sky brightness is inversely related to the distance between ships and the colony, with closer ships increasing sky brightness more.**

```{r Calculate Distance}
#Skomer coordinates
skomercoord <- c(lon = -5.281967525708255, lat = 51.73763792687346)

#Get coordinates of ships into separate vector
shipcoord <- cbind(darkdata$lon, darkdata$lat)

#Replicate skomer coordinates for the number of ships we have positions for 
skomercoord <- matrix(rep(skomercoord, nrow(darkdata)), ncol = 2, byrow = TRUE) 

#Calculate distance in meters
darkdata$distance.metres <- distVincentyEllipsoid(shipcoord, skomercoord)

#Average distance of all the ships per time point
darkdata2 <- darkdata  %>%
  group_by(POSIXct) %>%
  summarise(
    distance.metres = median(distance.metres, na.rm = TRUE), 
    across(c(cdm2, days_from_start, hours_since_midnight), first),
    .groups = "drop")
```

```{r Model Creation and Output}
#Model
dismodel <- glmmTMB(
  cdm2 ~ distance.metres + days_from_start + hours_since_midnight + (1 | days_from_start),
  family = Gamma(link = "log"),
  data = darkdata2)

#Model output
summary(dismodel)
pred2 <- ggpredict(dismodel, terms = "distance.metres")
plot(pred2)

#Plot the model
distance <- ggplot() +
  geom_point(data = darkdata2,
             aes(x = distance.metres, y = cdm2),
             size = 1.5) +
  geom_line(data = pred2,
            aes(x = x, y = predicted),
            color = "blue", linewidth = 1) +
  geom_ribbon(data = pred2,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  scale_x_continuous(
    name = "Distance from study colony (m)",
    breaks = seq(0, max(darkdata2$distance.metres, na.rm = TRUE), by = 2500)) +
  scale_y_continuous(
    name = expression("Sky Brightness (" * cd/m^2 * ")")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 12), 
    axis.title.x = element_text(margin = margin(t = 14)), 
    axis.title.y = element_text(margin = margin(r = 14)))


```

## **Hypothesis 3**

**Sky brightness increases when ship lighting is oriented toward the colony compared to when it is oriented away.**

#### (1) Heading

```{r Calculate heading with respect to Skomer}
#Skomer coordinates (need to be separated this time)
skomerlon <-  -5.281967525708255    
skomerlat <- - 51.73763792687346 

########### Calculate bearing from each ship to skomer colony
#Geosphere package expects (lon,lat) pairs; returns degrees (may be negative)
bearing_to_sk <- bearing(
  p1 = as.matrix(darkdata[, c("lon","lat")]), 
  p2 = matrix(c(skomerlon, skomerlat), nrow = nrow(darkdata), ncol = 2, byrow = TRUE))

#Normalize bearings to 0 and 360
bearing_to_sk <- (bearing_to_sk + 360) %% 360

#Add bearing to dataframe
darkdata$bearing_to_sk <- bearing_to_sk

########## Compute relative heading
#We want: 0 = ship is heading directly toward Skomer
#Relative = bearing_to_sk - ship_heading
#Then convert to signed -180/+180
#Negative = need to turn left to face Skomer, positive = need to turn right to face Skomer 
rel_deg_0_360 <- (darkdata$bearing_to_sk - darkdata$heading + 360) %% 360
rel_deg_signed <- ifelse(rel_deg_0_360 > 180, rel_deg_0_360 - 360, rel_deg_0_360)

#Add to dataframe
darkdata$relative_heading_deg <- rel_deg_signed

#Interprete the heading number so its easier to understand
interpret <- function(x){
  if(is.na(x)) return(NA)
  if(x == 0) return("_Heading directly toward Skomer")
  if(abs(x) == 180) return("_Heading directly away from Skomer")
  if(x > 0) return(paste0(round(x,1),"_clockwise (turn right) to point at Skomer"))
  if(x < 0) return(paste0(round(abs(x),1),"_anticlockwise (turn left) to point at Skomer"))
}
darkdata$interpretation <- sapply(darkdata$relative_heading_deg, interpret)

#Split up interpretation from the heading number so we can plot heading 
darkdata <- separate(
  darkdata,
  interpretation,
  into = c("heading_sk", "interpretation"),
  sep = "_",
  remove = FALSE)

#Make sure it is numeric 
darkdata$heading_sk <- as.numeric(darkdata$heading_sk)

#Average heading of all the ships per time point
darkdata2 <- darkdata  %>%
  group_by(POSIXct) %>%
  summarise(
    heading_sk = median(heading_sk, na.rm = TRUE), 
    across(c(cdm2, days_from_start, hours_since_midnight), first),
    .groups = "drop")
```

```{r Model Creation and Output}
#Model
headmodel <- glmmTMB(
  cdm2 ~ heading_sk + days_from_start + hours_since_midnight + (1 | days_from_start),
  family = Gamma(link = "log"), 
  data = darkdata2)

#Model output
summary(headmodel)
pred3 <- ggpredict(headmodel, terms = "heading_sk")
plot(pred3)

#Plot the model
head <- ggplot() +
  geom_point(data = darkdata2,
             aes(x = heading_sk, y = cdm2),
             size = 1.5) +
  geom_line(data = pred3,
            aes(x = x, y = predicted),
            color = "blue", linewidth = 1) +
  geom_ribbon(data = pred3,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  scale_x_continuous(
    name = "Heading (°)",
    breaks = seq(0, 180, by = 30)) +
  scale_y_continuous(
    name = expression("Sky Brightness (" * cd/m^2 * ")")) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 12), 
    axis.title.x = element_text(margin = margin(t = 14)), 
    axis.title.y = element_text(margin = margin(r = 14)))
```

#### (2) Orientation

```{r Model Creation and Output}
#Define heading as an orientation 
darkdata2$heading_bin <- cut(
  darkdata2$heading_sk,
  breaks = c(0, 45, 135, 180),
  labels = c("Front", "Side", "Rear"))
darkdata2$heading_bin <- factor(darkdata2$heading_bin)

#Model 
headgroup <- glmmTMB(
  cdm2 ~ heading_bin + hours_since_midnight + (1 | days_from_start),
  family = Gamma(link = "log"),
  data = darkdata2)

#Model Output
summary(headgroup)

#Pairwise comparisons
emm <- emmeans(headgroup, ~ heading_bin)
pw <- as.data.frame(pairs(emm, adjust = "tukey"))

#Plot the model
#Split "A - B" into two columns for plotting brackets
pw <- pw %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    p.signif = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      TRUE ~ "ns"))

#Set bracket heights
ymax <- max(darkdata2$cdm2, na.rm = TRUE)
yrng <- diff(range(darkdata2$cdm2, na.rm = TRUE))
pw$y.position <- ymax + seq(0.08, by = 0.08, length.out = nrow(pw)) * yrng

headgroup <- ggplot(darkdata2, aes(x = heading_bin, y = cdm2, fill = heading_bin)) + 
  geom_boxplot(width = 0.65, alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.12, size = 1.0) +
  scale_fill_brewer(palette = "Set2") +
  stat_pvalue_manual(
    pw,
    label = "p.signif",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    hide.ns = TRUE,
    size = 7
  ) +
  scale_y_continuous(
    name = expression("Sky Brightness (" * cd/m^2 * ")")
  ) +
  scale_x_discrete(
    name = "Heading Orientation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 15),
    axis.text  = element_text(size = 14),
    axis.title.x = element_text(margin = margin(t = 14)),
    axis.title.y = element_text(margin = margin(r = 14)),
    legend.position = "none")
```

```{r Final Plot}
#Plot all 4 together with lettering 
(ship | distance) /
  (head | headgroup) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```

## **Is Artificial Light at Night able to replicate a Full Moon?** 

```{r Housekeeping}
####### Read in the light data
lightdata <- read.csv("~/Library/CloudStorage/OneDrive-Nexus365/4th Year 2025-2026/Analysis/analysis/data/Light.csv")

####### Split data into moon cyles
#Get moon illumination info
moon <- getMoonIllumination(date = lightdata$POSIXct)

#Create phase names by the phase percentage
phase_name <- function(p) {
  if (p < 0.0625 || p >= 0.9375) return("New Moon")
  if (p < 0.1875) return("Waxing Crescent")
  if (p < 0.3125) return("First Quarter")
  if (p < 0.4375) return("Waxing Gibbous")
  if (p < 0.5625) return("Full Moon")
  if (p < 0.6875) return("Waning Gibbous")
  if (p < 0.8125) return("Last Quarter")
  return("Waning Crescent")}

lightdata$moonPhaseName <- vapply(moon$phase, phase_name, character(1))

######## Night of ID (same as above)
sun <-getSunlightTimes(date = as.Date(lightdata$POSIXct), tz = "UTC", lat= 51.737449, lon=-5.287829, keep =c("sunrise",  "sunset"))
lightdata <- cbind(lightdata, sun)
lightdata <- subset(lightdata, select = -c(date, lat, lon))
#Give night.id again (same as before)
lightdata$night.of <- with(lightdata, as.Date(
  ifelse(POSIXct >= sunset, as.Date(sunset),           
         ifelse(POSIXct < sunrise, as.Date(sunrise) - 1, as.Date(POSIXct)))))

####### Filtering
#Remove data when sun is having an effect (inflates and we only want to look at lunar effects)
lightdata <- filter(lightdata, twilightModel < 0.00001)

#Filter for dark (ALAN only)
lightdata$illumination <- as.numeric(lightdata$illumination)
dark <- filter(lightdata, illumination == 0)

#Remove data points when moon is below horizon (only want to look at when moon is actually up - dont want to compare ALAN with a night when moon can reach 50% but has set already)
lunar <- filter(lightdata, illumination > 0)

####### Merging datasets 
#Assign ALAN a category name so I can merge with lunar phase data
dark <- dark %>%
  mutate(moonPhaseName = "ALAN")
combined <- bind_rows(lunar, dark)

####### Get a mean (mean, median, maximum) for each lunar phase and for ALAN
combmean <- combined %>%
  group_by(moonPhaseName) %>%
  summarise(
    mean_cdm2 = mean(cdm2, na.rm = TRUE),
    mean_logcdm2 = mean(logcdm2, na.rm = TRUE),
    max_cdm2 = max(cdm2, na.rm = TRUE), 
    max_logcdm2 = max(logcdm2, na.rm = TRUE),
    med_cdm2 = median(cdm2, na.rm = TRUE),
    med_logcdm2 = median(logcdm2, na.rm = TRUE),
    .groups = "drop")

#Look at cd/m2 for ALAN and all the moon phases  
max(dark$logcdm2)
max(lunar$logcdm2)
mean(dark$logcdm2)
mean(lunar$logcdm2)

####### Housekeeping for plotting
#Order labels so they plot in moon cycle order
combined$moonPhaseName <- factor(
  combined$moonPhaseName,
  levels = rev(c(
    "New Moon", "Waxing Crescent", "First Quarter",
    "Waxing Gibbous", "Full Moon",
    "Waning Gibbous", "Last Quarter",
    "Waning Crescent",
    "ALAN")))

#Colours based on how bright it is 
phase_cols <- c(
  "New Moon" = "#1750AC",
  "Waxing Crescent" = "#3373C4",
  "First Quarter" = "#73B9EE",
  "Waxing Gibbous" = "#E8C766",
  "Full Moon" = "#E8AF30",
  "Waning Gibbous" = "#E8C766",
  "Last Quarter" = "#73B9EE",
  "Waning Crescent" = "#3373C4", 
  "ALAN" = "#eb7d31")
```

```{r Model}
#Model to show difference in cd/m2 between ALAN and different moon phases
moongroup <- glmmTMB(logcdm2 ~ moonPhaseName, family = gaussian(), data = combined)

#Model output 
summary(moongroup)

#Pairwise comparisons from the fitted model
library(emmeans)
emm <- emmeans(moongroup, ~ moonPhaseName)
pw <- as.data.frame(pairs(emm, adjust = "tukey"))
```

```{r Plots}
#Boxplot
box <- ggplot(combined, aes(x = logcdm2, y = moonPhaseName, fill = moonPhaseName)) +
  geom_point(aes(color = moonPhaseName),
             position = position_jitter(width = 0.01),
             alpha = 0.8) +
  geom_boxplot(outlier.shape = NA, width = 0.7) +
  scale_fill_manual(values = phase_cols) +
  scale_color_manual(values = phase_cols) +
  scale_y_discrete(name = "Moon Phase", expand = expansion(add = 0.6)) +
  scale_x_continuous(name = expression("Sky Brightness (" * log (cd/m^2) * ")")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_markdown(size = 12, angle = 0, vjust = 0.8),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 15,margin = margin (r=15)),
    axis.title.x = element_text(size = 15, margin = margin (t=15)),
    plot.margin = margin(l = 10, b = 10, r = 10, t = 10))
#This will be an underestimation of ALAN bc the nat contains ALAN as well. But is the best we can do considering 

#Plot bar graph 
bar <- ggplot(combmean, aes(x = moonPhaseName, y = max_logcdm2, fill = moonPhaseName)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = phase_cols) +
  geom_hline(yintercept = 1.849424, linetype = "dashed", color = "#eb7d31")+
  scale_x_discrete(name = "Moon Phase", breaks = combmean$moonPhaseName, expand = expansion(add = 0.2)) +
  scale_y_continuous(name = expression("Maximum Sky Brightness (" * log (cd/m^2) * ")")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12),
    axis.text.y = element_markdown(size = 12),
    axis.title.x = element_text(size = 15,  margin = margin (t=15)),
    axis.title.y = element_text(size = 15, margin = margin (r=15)),
    plot.margin = margin(l = 10, b = 10, r = 10, t = 10))

#Plot all 2 together with lettering 
(box) /
  (bar) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```
